---
title: "Spatial Weights and Applications"
subtitle: "Hands-on Exercise 4"
author: "Zhiyi Hao"
date: "2026-09-16"
format:
  html:
    toc: true
    toc-depth: 4
    code-fold: false
    number-offset: 7
    fig-width: 8
    fig-height: 8
execute:
  echo: true
  warning: false
  message: false
---

# Spatial Weights and Applications {.course-chapter-title}

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 8,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)

library(sf)
library(tidyverse)
library(tmap)
library(spdep)
library(knitr)

tmap_mode("plot")
```

## Overview

This exercise uses the 88 county-level units in Hunan to define who is considered a neighbour. I use GDP per capita as the example indicator and follow the workflow from importing the data to calculating spatially lagged values.

The main questions I keep in mind are:

- How does the definition of a neighbour change the weight matrix?
- How does a county's surrounding GDP per capita compare with its own value?
- What changes when the focal county is included in its own spatial window?

## The Study Area and Data

The analysis uses the Hunan county boundaries and the 2012 indicator table supplied in `data4.zip`. The two files join through the county name.

### Getting Started

```{r}
#| label: ch08-01
pacman::p_load(sf, spdep, tmap, tidyverse, knitr)
```

## Getting the Data Into R Environment

### Import shapefile into r environment

The import paths below point to this exercise's `data` folder. The course paths are shown immediately before each local import.

```r
hunan <- st_read(dsn = "chap08/data/geospatial", 
                 layer = "Hunan")
```

```{r}
#| label: ch08-02
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan",
                 quiet = TRUE)
```

### Import csv file into r environment

```r
hunan2012 <- read_csv("chap08/data/aspatial/Hunan_2012.csv")
```

```{r}
#| label: ch08-03
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv", show_col_types = FALSE)
```

### Performing relational join

```{r}
#| label: ch08-04
hunan <- left_join(hunan,hunan2012) %>%
  select(1:4, 7, 15)
```

::: {.callout-tip}
### What I noticed {.unnumbered}

The join keeps one row for each county and adds `GDPPC` to the spatial layer. The geometry is still in WGS 84, which is suitable for this first map. The distance-based examples use longitude and latitude with `longlat = TRUE`, so the distances are in kilometres. I compare these with projected distances in my additional check.
:::

## Visualising Regional Development Indicator

```{r}
#| label: ch08-05
#| fig-width: 12
#| fig-height: 8
basemap <- tm_shape(hunan) +
  tm_polygons() +
  tm_text("NAME_3", size=0.5)

gdppc <- qtm(hunan, "GDPPC") +
  tm_layout(
    legend.position = c("left", "bottom"))
tmap_arrange(basemap, gdppc, asp=1, ncol=2)
```

::: {.callout-tip}
### My observation {.unnumbered}

The highest GDPPC values are concentrated around a small number of counties, while many western and southern counties sit in lower classes. The map gives the variable a spatial shape, yet it does not tell me whether neighbouring counties are similar enough to support a clustering claim. That is the reason for constructing weights before calculating autocorrelation statistics.
:::

## Computing Contiguity Spatial Weights

### Computing (QUEEN) contiguity based neighbours

```{r}
#| label: ch08-06
wm_q <- poly2nb(hunan, queen=TRUE)
summary(wm_q)
```

```{r}
#| label: ch08-07
wm_q[[1]]
```

```{r}
#| label: ch08-08
hunan$County[1]
```

```{r}
#| label: ch08-09
hunan$NAME_3[c(2,3,4,57,85)]
```

```{r}
#| label: ch08-10
nb1 <- wm_q[[1]]
nb1 <- hunan$GDPPC[nb1]
nb1
```

```{r}
#| label: ch08-11
str(wm_q)
```

### Creating (ROOK) contiguity based neighbours

```{r}
#| label: ch08-12
wm_r <- poly2nb(hunan, queen=FALSE)
summary(wm_r)
```

::: {.callout-tip}
### What I noticed {.unnumbered}

Queen contiguity counts a shared corner as a connection. Rook contiguity requires a shared boundary segment, so it usually produces fewer links. The choice changes the question: Queen captures broader contact, while Rook focuses on direct edge-sharing neighbours.
:::

### Visualising contiguity weights

```{r}
#| label: ch08-13
longitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[1]])
```

```{r}
#| label: ch08-14
latitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[2]])
```

```{r}
#| label: ch08-15
coords <- cbind(longitude, latitude)
```

```{r}
#| label: ch08-16
head(coords)
```

#### Plotting Queen contiguity based neighbours map

Each black point represents a county centroid. The red segments connect Queen neighbours.

```{r}
#| label: ch08-17
plot(hunan$geometry, border="lightgrey")
plot(wm_q, coords, pch = 19, cex = 0.6, add = TRUE, col= "red")
```

#### Plotting Rook contiguity based neighbours map

This map uses the same coordinates and red links with the Rook neighbour definition.

```{r}
#| label: ch08-18
plot(hunan$geometry, border="lightgrey")
plot(wm_r, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```

#### Plotting both Queen and Rook contiguity based neighbours maps

```{r}
#| label: ch08-19
#| fig-width: 12
#| fig-height: 8
par(mfrow=c(1,2))
plot(hunan$geometry, border="lightgrey", main="Queen Contiguity")
plot(wm_q, coords, pch = 19, cex = 0.6, add = TRUE, col= "red")
plot(hunan$geometry, border="lightgrey", main="Rook Contiguity")
plot(wm_r, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```

## Computing distance based neighbours

### Determine the cut-off distance

```{r}
#| label: ch08-20
#coords <- coordinates(hunan)
k1 <- knn2nb(knearneigh(coords))
k1dists <- unlist(nbdists(k1, coords, longlat = TRUE))
summary(k1dists)
```

### Computing fixed distance weight matrix

```{r}
#| label: ch08-21
wm_d62 <- dnearneigh(coords, 0, 62, longlat = TRUE)
wm_d62
```

```{r}
#| label: ch08-22
str(wm_d62)
```

```{r}
#| label: ch08-23
table(hunan$County, card(wm_d62))
```

```{r}
#| label: ch08-24
n_comp <- n.comp.nb(wm_d62)
n_comp$nc
```

```{r}
#| label: ch08-25
table(n_comp$comp.id)
```

#### Plotting fixed distance weight matrix

```{r}
#| label: ch08-26
plot(hunan$geometry, border="lightgrey")
plot(wm_d62, coords, add=TRUE)
plot(k1, coords, add=TRUE, col="red", length=0.08)
```

```{r}
#| label: ch08-27
#| fig-width: 12
#| fig-height: 8
par(mfrow=c(1,2))
plot(hunan$geometry, border="lightgrey", main="1st nearest neighbours")
plot(k1, coords, add=TRUE, col="red", length=0.08)
plot(hunan$geometry, border="lightgrey", main="Distance link")
plot(wm_d62, coords, add=TRUE, pch = 19, cex = 0.6)
```

### Computing adaptive distance weight matrix

```{r}
#| label: ch08-28
knn6 <- knn2nb(knearneigh(coords, k=6))
knn6
```

```{r}
#| label: ch08-29
str(knn6)
```

#### Plotting distance based neighbours

```{r}
#| label: ch08-30
plot(hunan$geometry, border="lightgrey")
plot(knn6, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```

::: {.callout-tip}
### My observation {.unnumbered}

The distance rule uses one global radius, so dense and sparse parts of the province can receive very different numbers of neighbours. The six-nearest-neighbour rule keeps the neighbour count stable and gives every county a comparable local sample. It changes the meaning of proximity from a fixed geography to a fixed amount of information.
:::

## Weights based on IDW

These weights use inverse great-circle distance in kilometres between Queen neighbours.

```{r}
#| label: ch08-31
dist <- nbdists(wm_q, coords, longlat = TRUE)
ids <- lapply(dist, function(x) 1/(x))
ids
```

## Row-standardised Weights Matrix

`style = "W"` normalises each row. The later `style = "B"` call preserves the supplied inverse-distance values.

```{r}
#| label: ch08-32
rswm_q <- nb2listw(wm_q, style="W", zero.policy = TRUE)
rswm_q
```

```{r}
#| label: ch08-33
rswm_q$weights[10]
```

```{r}
#| label: ch08-34
rswm_ids <- nb2listw(wm_q, glist=ids, style="B", zero.policy=TRUE)
rswm_ids
```

```{r}
#| label: ch08-35
rswm_ids$weights[1]
```

```{r}
#| label: ch08-36
summary(unlist(rswm_ids$weights))
```

::: {.callout-tip}
### What I noticed {.unnumbered}

Row-standardisation makes each county's neighbour contributions sum to one. A county with five neighbours gives each one-fifth of the total weight, while a county with two neighbours gives each one-half. This creates an average-like spatial lag and prevents counties with many neighbours from automatically dominating the calculation.
:::

## Application of Spatial Weight Matrix

### Spatial lag with row-standardized weights

```{r}
#| label: ch08-37
GDPPC.lag <- lag.listw(rswm_q, hunan$GDPPC)
GDPPC.lag
```

```{r}
#| label: ch08-38
nb1 <- wm_q[[1]]
nb1 <- hunan$GDPPC[nb1]
nb1
```

```{r}
#| label: ch08-39
lag.list <- list(hunan$NAME_3, lag.listw(rswm_q, hunan$GDPPC))
lag.res <- as.data.frame(lag.list)
colnames(lag.res) <- c("NAME_3", "lag GDPPC")
hunan <- left_join(hunan,lag.res)
```

```{r}
#| label: ch08-40
head(hunan)
```

```{r}
#| label: ch08-41
#| fig-width: 12
#| fig-height: 8
gdppc <- qtm(hunan, "GDPPC")
lag_gdppc <- qtm(hunan, "lag GDPPC")
tmap_arrange(gdppc, lag_gdppc, asp=1, ncol=2)
```

### Spatial lag as a sum of neighboring values

```{r}
#| label: ch08-42
b_weights <- lapply(wm_q, function(x) 0*x + 1)
b_weights2 <- nb2listw(wm_q, 
                       glist = b_weights, 
                       style = "B")
b_weights2
```

```{r}
#| label: ch08-43
lag_sum <- list(hunan$NAME_3, lag.listw(b_weights2, hunan$GDPPC))
lag.res <- as.data.frame(lag_sum)
colnames(lag.res) <- c("NAME_3", "lag_sum GDPPC")
```

```{r}
#| label: ch08-44
lag_sum
```

```{r}
#| label: ch08-45
hunan <- left_join(hunan, lag.res)
```

```{r}
#| label: ch08-46
#| fig-width: 12
#| fig-height: 8
gdppc <- qtm(hunan, "GDPPC")
lag_sum_gdppc <- qtm(hunan, "lag_sum GDPPC")
tmap_arrange(gdppc, lag_sum_gdppc, asp=1, ncol=2)
```

### Spatial window average

```{r}
#| label: ch08-47
wm_qs <- include.self(wm_q)
```

```{r}
#| label: ch08-48
wm_qs[[1]]
```

```{r}
#| label: ch08-49
wm_qs <- nb2listw(wm_qs)
wm_qs
```

```{r}
#| label: ch08-50
lag_w_avg_gpdpc <- lag.listw(wm_qs, 
                             hunan$GDPPC)
lag_w_avg_gpdpc
```

```{r}
#| label: ch08-51
lag.list.wm_qs <- list(hunan$NAME_3, lag.listw(wm_qs, hunan$GDPPC))
lag_wm_qs.res <- as.data.frame(lag.list.wm_qs)
colnames(lag_wm_qs.res) <- c("NAME_3", "lag_window_avg GDPPC")
```

```{r}
#| label: ch08-52
hunan <- left_join(hunan, lag_wm_qs.res)
```

```{r}
#| label: ch08-53
hunan %>%
  select("County", 
         "lag GDPPC", 
         "lag_window_avg GDPPC") %>%
  kable()
```

```{r}
#| label: ch08-54
#| fig-width: 12
#| fig-height: 8
w_avg_gdppc <- qtm(hunan, "lag_window_avg GDPPC")
tmap_arrange(lag_gdppc, w_avg_gdppc, asp=1, ncol=2)
```

### Spatial window sum

```{r}
#| label: ch08-55
wm_qs <- include.self(wm_q)
wm_qs
```

```{r}
#| label: ch08-56
b_weights <- lapply(wm_qs, function(x) 0*x + 1)
b_weights[1]
```

```{r}
#| label: ch08-57
b_weights2 <- nb2listw(wm_qs, 
                       glist = b_weights, 
                       style = "B")
b_weights2
```

```{r}
#| label: ch08-58
w_sum_gdppc <- list(hunan$NAME_3, lag.listw(b_weights2, hunan$GDPPC))
w_sum_gdppc
```

```{r}
#| label: ch08-59
w_sum_gdppc.res <- as.data.frame(w_sum_gdppc)
colnames(w_sum_gdppc.res) <- c("NAME_3", "w_sum GDPPC")
```

```{r}
#| label: ch08-60
hunan <- left_join(hunan, w_sum_gdppc.res)
```

```{r}
#| label: ch08-61
hunan %>%
  select("County", "lag_sum GDPPC", "w_sum GDPPC") %>%
  kable()
```

```{r}
#| label: ch08-62
#| fig-width: 12
#| fig-height: 8
w_sum_gdppc <- qtm(hunan, "w_sum GDPPC")
tmap_arrange(lag_sum_gdppc, w_sum_gdppc, asp=1, ncol=2)
```

::: {.callout-tip}
### My observation {.unnumbered}

The row-standardised lag is comparable across counties because it averages the neighbours. The binary lag sum is larger in places with more neighbours, so it measures a sum of neighbouring per-capita indicators. This sum is not total GDP, which would require population weights. Adding the focal county to the window moves the result closer to a local context measure that includes both the county and its immediate surroundings.
:::

## Final Takeaway {.unnumbered}


The weight matrix is part of the analysis question. Queen and Rook describe different forms of contact, fixed-distance and k-nearest-neighbour rules describe different forms of proximity, and row-standardised or binary weights produce different meanings for a spatial lag.

For this Hunan example, GDPPC appears geographically uneven in the map and the neighbouring values are structured enough to justify checking spatial autocorrelation in the next exercise. I would report the selected neighbour rule with the statistic because the result is not independent of that modelling choice.

::: {.callout-tip}
## Beyond the Code: The Weight Matrix Is an Assumption {.unnumbered}

I initially treated a neighbour as a simple yes-or-no relationship. The calculations made the assumption more visible: the same GDPPC values can produce different lag surfaces when I change Queen to Rook, use a distance threshold, or standardise each row. A spatial statistic is therefore not only a property of the data. It is also a property of how local context is defined.
:::



## References

- Course source: [Spatial Weights and Applications](https://r4gdsa.netlify.app/chap08.html)
- `spdep` documentation: [Creating neighbours using sf objects](https://cran.r-project.org/web/packages/spdep/vignettes/nb_sf.html)
