---
title: "Spatial Weights and Applications"
subtitle: "Hands-on Exercise 4"
author: "Zhiyi Hao"
date: "2026-09-16"
format:
  html:
    toc: true
    toc-depth: 4
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 6,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)

library(sf)
library(tidyverse)
library(tmap)
library(spdep)
library(knitr)

tmap_mode("plot")
```

## Overview

This exercise uses the 88 county-level units in Hunan to define who is considered a neighbour. I use GDP per capita as the example indicator and follow the workflow from importing the data to calculating spatially lagged values.

The main questions I keep in mind are:

- How does the definition of a neighbour change the weight matrix?
- How does a county's surrounding GDP per capita compare with its own value?
- What changes when the focal county is included in its own spatial window?

::: {.callout-note}
### Local data note

The supplied `data4.zip` contains the same Hunan boundary and 2012 indicator table used in the course example. The files are stored locally under `Hands-on_Ex04/data/`, so every rendered result on this page is generated from the repository rather than an absolute computer path.
:::

## Getting the Data Into R

### Loading packages

The original package-loading command is kept here for reference.

```{r}
#| label: course-packages
#| eval: false
pacman::p_load(sf, spdep, tmap, tidyverse, knitr)
```

```{r}
#| label: local-packages
library(sf)
library(spdep)
library(tmap)
library(tidyverse)
library(knitr)
```

### Importing the boundary

```{r}
#| label: course-read-shapefile
#| eval: false
hunan <- st_read(dsn = "chap08/data/geospatial", 
                 layer = "Hunan")
```

```{r}
#| label: local-read-shapefile
hunan <- st_read(
  dsn = "data/geospatial",
  layer = "Hunan",
  quiet = TRUE
)
hunan
```

### Importing the indicator table

```{r}
#| label: course-read-csv
#| eval: false
hunan2012 <- read_csv("chap08/data/aspatial/Hunan_2012.csv")
```

```{r}
#| label: local-read-csv
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv", show_col_types = FALSE)
```

### Joining the attributes

```{r}
#| label: course-relational-join
#| eval: false
hunan <- left_join(hunan,hunan2012) %>%
  select(1:4, 7, 15)
```

```{r}
#| label: local-relational-join
hunan <- left_join(hunan, hunan2012, by = "County") %>%
  select(1:4, 7, 15)

tibble(
  counties = nrow(hunan),
  indicator = "GDP per capita",
  missing_gdppc = sum(is.na(hunan$GDPPC)),
  source_crs = st_crs(hunan)$input
)
```

```{r}
#| label: course-head-hunan
#| eval: false
head(hunan)
```

```{r}
#| label: head-hunan
head(hunan) %>%
  st_drop_geometry() %>%
  select(County, GDPPC, NAME_3)
```

::: {.callout-tip}
### What I noticed

The join keeps one row for each county and adds `GDPPC` to the spatial layer. The geometry is still in WGS 84, which is suitable for this first map. I keep a projected copy for distance-based calculations later in the page.
:::

## Visualising Regional Development

```{r}
#| label: course-basemap
#| eval: false
basemap <- tm_shape(hunan) +
  tm_polygons() +
  tm_text("NAME_3", size=0.5)

gdppc <- qtm(hunan, "GDPPC") +
  tm_layout(
    legend.position = c("left", "bottom"))
tmap_arrange(basemap, gdppc, asp=1, ncol=2)
```

```{r}
#| label: regional-development-map
#| fig-cap: "County boundaries and GDP per capita across Hunan in 2012."
tm_shape(hunan) +
  tm_polygons(
    fill = "GDPPC",
    fill.scale = tm_scale_intervals(n = 6, style = "quantile"),
    fill.legend = tm_legend(title = "GDPPC")
  ) +
  tm_text("NAME_3", size = 0.45) +
  tm_layout(frame = FALSE)
```

::: {.callout-tip}
### My observation

The highest GDPPC values are concentrated around a small number of counties, while many western and southern counties sit in lower classes. The map gives the variable a spatial shape, yet it does not tell me whether neighbouring counties are similar enough to support a clustering claim. That is the reason for constructing weights before calculating autocorrelation statistics.
:::

## Contiguity Spatial Weights

### Queen neighbours

```{r}
#| label: course-queen-neighbours
#| eval: false
wm_q <- poly2nb(hunan, queen=TRUE)
summary(wm_q)
```

```{r}
#| label: queen-neighbour-summary
wm_q <- poly2nb(hunan, queen = TRUE)
queen_summary <- summary(wm_q)
queen_summary
```

```{r}
#| label: course-first-neighbour-list
#| eval: false
wm_q[[1]]
```

```{r}
#| label: first-neighbour-list
wm_q[[1]]
```

```{r}
#| label: course-first-county
#| eval: false
hunan$County[1]
```

```{r}
#| label: first-county
hunan$County[1]
```

```{r}
#| label: course-neighbour-names
#| eval: false
hunan$NAME_3[c(2,3,4,57,85)]
```

```{r}
#| label: neighbour-names
hunan$NAME_3[wm_q[[1]]]
```

```{r}
#| label: course-neighbour-gdppc
#| eval: false
nb1 <- wm_q[[1]]
nb1 <- hunan$GDPPC[nb1]
nb1
```

```{r}
#| label: neighbour-gdppc
queen_gdppc <- tibble(
  county = hunan$NAME_3[wm_q[[1]]],
  GDPPC = hunan$GDPPC[wm_q[[1]]]
)
queen_gdppc
```

```{r}
#| label: course-queen-structure
#| eval: false
str(wm_q)
```

### Rook neighbours

```{r}
#| label: course-rook-neighbours
#| eval: false
wm_r <- poly2nb(hunan, queen=FALSE)
summary(wm_r)
```

```{r}
#| label: rook-neighbour-summary
wm_r <- poly2nb(hunan, queen = FALSE)
rook_summary <- summary(wm_r)
rook_summary
```

::: {.callout-tip}
### What I noticed

Queen contiguity counts a shared corner as a connection. Rook contiguity requires a shared boundary segment, so it usually produces fewer links. The choice changes the question: Queen captures broader contact, while Rook focuses on direct edge-sharing neighbours.
:::

### Visualising contiguity

```{r}
#| label: course-centroid-longitude
#| eval: false
longitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[1]])
```

```{r}
#| label: course-centroid-latitude
#| eval: false
latitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[2]])
```

```{r}
#| label: course-centroid-coordinates
#| eval: false
coords <- cbind(longitude, latitude)
head(coords)
```

```{r}
#| label: local-centroid-coordinates
centroids <- st_centroid(hunan)
coords <- st_coordinates(centroids)
head(coords)
```

```{r}
#| label: course-queen-plot
#| eval: false
plot(hunan$geometry, border="lightgrey")
plot(wm_q, coords, pch = 19, cex = 0.6, add = TRUE, col= "red")
```

```{r}
#| label: course-rook-plot
#| eval: false
plot(hunan$geometry, border="lightgrey")
plot(wm_r, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```

```{r}
#| label: contiguity-comparison
#| fig-cap: "Queen and Rook contiguity links drawn between county centroids."
par(mfrow = c(1, 2))
plot(hunan$geometry, border = "lightgrey", main = "Queen Contiguity")
plot(wm_q, coords, pch = 19, cex = 0.6, add = TRUE, col = "#7fae15")
plot(hunan$geometry, border = "lightgrey", main = "Rook Contiguity")
plot(wm_r, coords, pch = 19, cex = 0.6, add = TRUE, col = "#2b8c76")
par(mfrow = c(1, 1))
```

```{r}
#| label: course-queen-rook-comparison
#| eval: false
par(mfrow=c(1,2))
plot(hunan$geometry, border="lightgrey", main="Queen Contiguity")
plot(wm_q, coords, pch = 19, cex = 0.6, add = TRUE, col= "red")
plot(hunan$geometry, border="lightgrey", main="Rook Contiguity")
plot(wm_r, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```

## Distance-Based Neighbours

### Checking the cut-off

The original exercise checks the nearest-neighbour distances with geographic coordinates.

```{r}
#| label: course-nearest-distance
#| eval: false
#coords <- coordinates(hunan)
k1 <- knn2nb(knearneigh(coords))
k1dists <- unlist(nbdists(k1, coords, longlat = TRUE))
summary(k1dists)
```

```{r}
#| label: projected-distance-preparation
hunan_m <- st_transform(hunan, 32650)
coords_m <- st_coordinates(st_centroid(hunan_m))
k1_m <- knn2nb(knearneigh(coords_m))
k1dists_m <- unlist(nbdists(k1_m, coords_m, longlat = FALSE))
summary(k1dists_m)
```

```{r}
#| label: course-distance-neighbours
#| eval: false
wm_d62 <- dnearneigh(coords, 0, 62, longlat = TRUE)
wm_d62
```

```{r}
#| label: projected-distance-neighbours
wm_d62 <- dnearneigh(coords_m, 0, 62000, longlat = FALSE)
wm_d62
```

```{r}
#| label: course-distance-structure
#| eval: false
str(wm_d62)
```

```{r}
#| label: distance-neighbour-counts
distance_counts <- tibble(
  county = hunan$NAME_3,
  neighbours = card(wm_d62)
)
distance_counts %>%
  summarise(
    counties = n(),
    min_neighbours = min(neighbours),
    max_neighbours = max(neighbours),
    mean_neighbours = mean(neighbours),
    disconnected = sum(neighbours == 0)
  )
```

```{r}
#| label: course-distance-frequency
#| eval: false
table(hunan$County, card(wm_d62))
```

```{r}
#| label: distance-frequency
distance_frequency <- as.data.frame(table(card(wm_d62)))
names(distance_frequency) <- c("neighbours", "counties")
distance_frequency
```

```{r}
#| label: course-component-check
#| eval: false
n_comp <- n.comp.nb(wm_d62)
n_comp$nc
table(n_comp$comp.id)
```

```{r}
#| label: component-check
n_comp <- n.comp.nb(wm_d62)
tibble(
  components = n_comp$nc,
  largest_component = max(table(n_comp$comp.id)),
  smallest_component = min(table(n_comp$comp.id))
)
```

```{r}
#| label: course-distance-plot
#| eval: false
plot(hunan$geometry, border="lightgrey")
plot(wm_d62, coords, add=TRUE)
plot(k1, coords, add=TRUE, col="red", length=0.08)
```

```{r}
#| label: distance-neighbour-map
#| fig-cap: "Projected distance links within 62 km, with one-nearest-neighbour links highlighted."
plot(st_geometry(hunan_m), border = "#d5ded2", main = "Distance-based neighbours")
plot(wm_d62, coords_m, add = TRUE, col = "#8aae1b")
plot(k1_m, coords_m, add = TRUE, col = "#b54b4b", length = 0.08)
```

```{r}
#| label: course-distance-nearest-comparison
#| eval: false
par(mfrow=c(1,2))
plot(hunan$geometry, border="lightgrey", main="1st nearest neighbours")
plot(k1, coords, add=TRUE, col="red", length=0.08)
plot(hunan$geometry, border="lightgrey", main="Distance link")
plot(wm_d62, coords, add=TRUE, pch = 19, cex = 0.6)
```

### Adaptive neighbours

```{r}
#| label: course-six-neighbours
#| eval: false
knn6 <- knn2nb(knearneigh(coords, k=6))
knn6
str(knn6)
plot(hunan$geometry, border="lightgrey")
plot(knn6, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```

```{r}
#| label: six-neighbour-summary
knn6_m <- knn2nb(knearneigh(coords_m, k = 6))
tibble(
  counties = length(knn6_m),
  links = sum(card(knn6_m)),
  neighbours_per_county = unique(card(knn6_m))
)
```

::: {.callout-tip}
### My observation

The distance rule uses one global radius, so dense and sparse parts of the province can receive very different numbers of neighbours. The six-nearest-neighbour rule keeps the neighbour count stable and gives every county a comparable local sample. It changes the meaning of proximity from a fixed geography to a fixed amount of information.
:::

## Weights Based on Inverse Distance

```{r}
#| label: course-idw-distances
#| eval: false
dist <- nbdists(wm_q, coords, longlat = TRUE)
ids <- lapply(dist, function(x) 1/(x))
ids
```

```{r}
#| label: projected-idw-weights
dist_m <- nbdists(wm_q, coords_m, longlat = FALSE)
ids_m <- lapply(dist_m, function(x) 1 / x)
summary(unlist(ids_m))
```

## Row-Standardised Weights

```{r}
#| label: course-row-standardised-weights
#| eval: false
rswm_q <- nb2listw(wm_q, style="W", zero.policy = TRUE)
rswm_q
```

```{r}
#| label: row-standardised-weights
rswm_q <- nb2listw(wm_q, style = "W", zero.policy = TRUE)
rswm_q
```

```{r}
#| label: course-first-row-weights
#| eval: false
rswm_q$weights[10]
```

```{r}
#| label: first-row-weights
rswm_q$weights[[1]]
```

```{r}
#| label: course-idw-listw
#| eval: false
rswm_ids <- nb2listw(wm_q, glist=ids, style="B", zero.policy=TRUE)
rswm_ids
```

```{r}
#| label: projected-idw-listw
rswm_ids_m <- nb2listw(wm_q, glist = ids_m, style = "B", zero.policy = TRUE)
summary(unlist(rswm_ids_m$weights))
```

```{r}
#| label: course-idw-first-row
#| eval: false
rswm_ids$weights[1]
```

```{r}
#| label: course-idw-summary
#| eval: false
summary(unlist(rswm_ids$weights))
```

::: {.callout-tip}
### What I noticed

Row-standardisation makes each county's neighbour contributions sum to one. A county with five neighbours gives each one-fifth of the total weight, while a county with two neighbours gives each one-half. This creates an average-like spatial lag and prevents counties with many neighbours from automatically dominating the calculation.
:::

## Applying Spatial Weight Matrices

### Spatial lag with row-standardised weights

```{r}
#| label: course-row-lag
#| eval: false
GDPPC.lag <- lag.listw(rswm_q, hunan$GDPPC)
GDPPC.lag
```

```{r}
#| label: row-lag
hunan[["lag GDPPC"]] <- lag.listw(rswm_q, hunan$GDPPC)
hunan %>%
  st_drop_geometry() %>%
  select(County, GDPPC, `lag GDPPC`) %>%
  slice_head(n = 10)
```

```{r}
#| label: course-row-lag-table
#| eval: false
lag.list <- list(hunan$NAME_3, lag.listw(rswm_q, hunan$GDPPC))
lag.res <- as.data.frame(lag.list)
colnames(lag.res) <- c("NAME_3", "lag GDPPC")
hunan <- left_join(hunan,lag.res)
```

```{r}
#| label: row-lag-map
#| fig-cap: "GDPPC and the Queen row-standardised spatial lag."
tm_shape(hunan) +
  tm_polygons(fill = "lag GDPPC", fill.scale = tm_scale_intervals(n = 6, style = "quantile")) +
  tm_layout(frame = FALSE)
```

```{r}
#| label: course-row-lag-head
#| eval: false
head(hunan)
```

```{r}
#| label: course-row-lag-comparison
#| eval: false
gdppc <- qtm(hunan, "GDPPC")
lag_gdppc <- qtm(hunan, "lag GDPPC")
tmap_arrange(gdppc, lag_gdppc, asp=1, ncol=2)
```

### Spatial lag as a sum of neighbouring values

```{r}
#| label: course-binary-weights
#| eval: false
b_weights <- lapply(wm_q, function(x) 0*x + 1)
b_weights2 <- nb2listw(wm_q, 
                       glist = b_weights, 
                       style = "B")
b_weights2
```

```{r}
#| label: binary-weights
b_weights <- lapply(wm_q, function(x) 0 * x + 1)
b_weights2 <- nb2listw(wm_q, glist = b_weights, style = "B")
hunan[["lag_sum GDPPC"]] <- lag.listw(b_weights2, hunan$GDPPC)
hunan %>%
  st_drop_geometry() %>%
  select(County, GDPPC, `lag_sum GDPPC`) %>%
  slice_head(n = 10)
```

```{r}
#| label: course-binary-lag-table
#| eval: false
lag_sum <- list(hunan$NAME_3, lag.listw(b_weights2, hunan$GDPPC))
lag.res <- as.data.frame(lag_sum)
colnames(lag.res) <- c("NAME_3", "lag_sum GDPPC")
```

```{r}
#| label: binary-lag-map
#| fig-cap: "GDPPC and the binary spatial lag sum."
tm_shape(hunan) +
  tm_polygons(fill = "lag_sum GDPPC", fill.scale = tm_scale_intervals(n = 6, style = "quantile")) +
  tm_layout(frame = FALSE)
```

```{r}
#| label: course-binary-lag-output
#| eval: false
lag_sum
```

```{r}
#| label: course-binary-lag-join
#| eval: false
hunan <- left_join(hunan, lag.res)
```

```{r}
#| label: course-binary-lag-comparison
#| eval: false
gdppc <- qtm(hunan, "GDPPC")
lag_sum_gdppc <- qtm(hunan, "lag_sum GDPPC")
tmap_arrange(gdppc, lag_sum_gdppc, asp=1, ncol=2)
```

### Spatial window average

```{r}
#| label: course-self-neighbours
#| eval: false
wm_qs <- include.self(wm_q)
wm_qs[[1]]
wm_qs <- nb2listw(wm_qs)
wm_qs
```

```{r}
#| label: self-neighbour-average
wm_qs <- include.self(wm_q)
wm_qs_listw <- nb2listw(wm_qs, style = "W", zero.policy = TRUE)
hunan[["lag_window_avg GDPPC"]] <- lag.listw(wm_qs_listw, hunan$GDPPC)

hunan %>%
  st_drop_geometry() %>%
  select(County, `lag GDPPC`, `lag_window_avg GDPPC`) %>%
  slice_head(n = 10) %>%
  kable()
```

```{r}
#| label: course-window-average-output
#| eval: false
lag_w_avg_gpdpc <- lag.listw(wm_qs,
                           hunan$GDPPC)
lag_w_avg_gpdpc
```

```{r}
#| label: course-window-average-table-data
#| eval: false
lag.list.wm_qs <- list(hunan$NAME_3, lag.listw(wm_qs, hunan$GDPPC))
lag_wm_qs.res <- as.data.frame(lag.list.wm_qs)
colnames(lag_wm_qs.res) <- c("NAME_3", "lag_window_avg GDPPC")
```

```{r}
#| label: course-window-average-join
#| eval: false
hunan <- left_join(hunan, lag_wm_qs.res)
```

```{r}
#| label: course-window-average-table
#| eval: false
hunan %>%
  select("County",
         "lag GDPPC",
         "lag_window_avg GDPPC") %>%
  kable()
```

```{r}
#| label: course-window-average-comparison
#| eval: false
w_avg_gdppc <- qtm(hunan, "lag_window_avg GDPPC")
tmap_arrange(lag_gdppc, w_avg_gdppc, asp=1, ncol=2)
```

```{r}
#| label: window-average-map
#| fig-cap: "Row-standardised lag and spatial-window average."
tm_shape(hunan) +
  tm_polygons(fill = "lag_window_avg GDPPC", fill.scale = tm_scale_intervals(n = 6, style = "quantile")) +
  tm_layout(frame = FALSE)
```

### Spatial window sum

```{r}
#| label: course-window-sum-setup
#| eval: false
wm_qs <- include.self(wm_q)
wm_qs
```

```{r}
#| label: course-window-sum-binary
#| eval: false
b_weights <- lapply(wm_qs, function(x) 0*x + 1)
b_weights[1]
```

```{r}
#| label: course-window-sum-listw
#| eval: false
b_weights2 <- nb2listw(wm_qs,
                     glist = b_weights,
                     style = "B")
b_weights2
```

```{r}
#| label: course-window-sum
#| eval: false
wm_qs <- include.self(wm_q)
b_weights <- lapply(wm_qs, function(x) 0*x + 1)
b_weights2 <- nb2listw(wm_qs, 
                       glist = b_weights, 
                       style = "B")
w_sum_gdppc <- lag.listw(b_weights2, hunan$GDPPC)
```

```{r}
#| label: course-window-sum-output
#| eval: false
w_sum_gdppc <- list(hunan$NAME_3, lag.listw(b_weights2, hunan$GDPPC))
w_sum_gdppc
```

```{r}
#| label: course-window-sum-dataframe
#| eval: false
w_sum_gdppc.res <- as.data.frame(w_sum_gdppc)
colnames(w_sum_gdppc.res) <- c("NAME_3", "w_sum GDPPC")
```

```{r}
#| label: course-window-sum-join
#| eval: false
hunan <- left_join(hunan, w_sum_gdppc.res)
```

```{r}
#| label: window-sum
wm_qs <- include.self(wm_q)
self_binary <- lapply(wm_qs, function(x) 0 * x + 1)
self_binary_listw <- nb2listw(wm_qs, glist = self_binary, style = "B")
hunan[["w_sum GDPPC"]] <- lag.listw(self_binary_listw, hunan$GDPPC)

hunan %>%
  st_drop_geometry() %>%
  select(County, `lag_sum GDPPC`, `w_sum GDPPC`) %>%
  slice_head(n = 10) %>%
  kable()
```

```{r}
#| label: course-window-sum-table
#| eval: false
hunan %>%
  select("County", "lag_sum GDPPC", "w_sum GDPPC") %>%
  kable()
```

```{r}
#| label: window-sum-map
#| fig-cap: "Binary lag sum and spatial-window sum."
tm_shape(hunan) +
  tm_polygons(fill = "w_sum GDPPC", fill.scale = tm_scale_intervals(n = 6, style = "quantile")) +
  tm_layout(frame = FALSE)
```

```{r}
#| label: course-window-sum-comparison
#| eval: false
w_sum_gdppc <- qtm(hunan, "w_sum GDPPC")
tmap_arrange(lag_sum_gdppc, w_sum_gdppc, asp=1, ncol=2)
```

::: {.callout-tip}
### My observation

The row-standardised lag is comparable across counties because it averages the neighbours. The binary lag sum is larger in places with more neighbours, so it measures the total surrounding GDPPC rather than a typical neighbour. Adding the focal county to the window moves the result closer to a local context measure that includes both the county and its immediate surroundings.
:::

## Final Takeaway

The weight matrix is part of the analysis question. Queen and Rook describe different forms of contact, fixed-distance and k-nearest-neighbour rules describe different forms of proximity, and row-standardised or binary weights produce different meanings for a spatial lag.

For this Hunan example, GDPPC appears geographically uneven in the map and the neighbouring values are structured enough to justify checking spatial autocorrelation in the next exercise. I would report the selected neighbour rule with the statistic because the result is not independent of that modelling choice.

::: {.callout-tip}
## Beyond the Code: The Weight Matrix Is an Assumption

I initially treated a neighbour as a simple yes-or-no relationship. The calculations made the assumption more visible: the same GDPPC values can produce different lag surfaces when I change Queen to Rook, use a distance threshold, or standardise each row. A spatial statistic is therefore not only a property of the data. It is also a property of how local context is defined.
:::

## References

- Course source: [Spatial Weights and Applications](https://r4gdsa.netlify.app/chap08.html)
- `spdep` documentation: [Creating neighbours using sf objects](https://cran.r-project.org/web/packages/spdep/vignettes/nb_sf.html)
