---
title: "Geospatial Data Science with R"
subtitle: "Hands-on Exercise 1"
author: "Zhiyi Hao"
date: "2026-08-26"
execute:
  echo: true
  message: false
  warning: false
format:
  html:
    code-fold: false
    number-offset: 0
---

# Geospatial Data Science with R {.course-chapter-title}

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 6,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

## Learning Outcome

I use this page to reproduce the Chapter 1 workflow with the local data folder. My focus is on the basic geospatial steps: importing layers, checking CRS, converting tabular coordinates into `sf`, and doing simple geometry operations.

::: {.callout-note}
**Local implementation note.**

All data for this exercise is stored in `Hands-on_Ex/Hands-on_Ex01/data`, so the code below uses local relative paths such as `data/geospatial` and `data/aspatial/listings.csv`.
:::

## Data Acquisition

Before running the code, I found it useful to separate the data by spatial type because each type supports a different kind of question.

| Spatial data type | Dataset in this exercise | How I use it |
|---|---|---|
| Points | Preschool locations and Airbnb listings | Represent individual facilities or listings |
| Polygons | Planning subzones | Define administrative areas for counting and density |
| Lines | Cycling paths | Represent a network-like feature for buffering |

### Extracting the geospatial data sets

::: {.callout-note}
**My learning point.**

The same map can mix points, polygons, and lines, but the operations are not interchangeable. For example, `st_intersects()` answers a point-in-polygon question, while `st_buffer()` expands a line into an area that can be measured.
:::

### Extracting the aspatial data set

The listing table is kept in the local `data/aspatial` folder and is introduced again when it is converted into a spatial layer.

## Getting Started

```{r}
pacman::p_load(sf, tidyverse)
```

```{r}
#| label: course-chap01-packages
#| eval: false
pacman::p_load(sf, tidyverse)
```

## Importing Geospatial Data

The first step is to bring the Singapore planning subzone boundary, cycling path layer, and preschool point layer into R. The Master Plan 2014 boundary and cycling path files are read as vector layers, while the preschool layer is read from the local KML derived from the downloaded GeoJSON.

### Importing polygon feature data in shapefile format

```{r}
mpsz = st_read(dsn = "data/geospatial", 
               layer = "MP14_SUBZONE_WEB_PL",
               quiet = TRUE)
```

```{r}
#| label: course-chap01-read-mpsz
#| eval: false
mpsz = st_read(dsn = "chap01/data/geospatial",
                  layer = "MP14_SUBZONE_WEB_PL")
```

### Importing polyline feature data in shapefile form

```{r}
cyclingpath = st_read(dsn = "data/geospatial", 
                      layer = "CyclingPathGazette",
                      quiet = TRUE)
```

```{r}
#| label: course-chap01-read-cyclingpath
#| eval: false
cyclingpath = st_read(dsn = "chap01/data/geospatial",
                         layer = "CyclingPathGazette")
```

```{r}
cyclingpath <- st_set_crs(cyclingpath, 3414)
```

### Importing GIS data in kml format

```{r}
preschool = st_read("data/geospatial/PreSchoolsLocation.kml",
                    quiet = TRUE) %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)
```

```{r}
#| label: course-chap01-read-preschool
#| eval: false
preschool = st_read("chap01/data/geospatial/PreSchoolsLocation.kml") %>%
  st_transform(crs = 3414)
```

The geometry column tells us what spatial type is stored in each object. For the planning subzones, the geometry is polygon-based because each row represents an areal unit.

## Checking the Content of A Simple Feature Data Frame

### Working with st_geometry()

```{r}
st_geometry(mpsz)
```

### Working with glimpse()

```{r}
glimpse(mpsz)
```

### Working with head()

```{r}
head(mpsz, n=5)
```

## Plotting the Geospatial Data

The default `plot()` method is a quick diagnostic view. It is not a publication map, but it is useful for confirming that the geometries and selected attributes look sensible.

```{r}
plot(mpsz)
```

```{r}
plot(st_geometry(mpsz))
```

```{r}
plot(mpsz["PLN_AREA_N"])
```

Overlaying preschool locations on the subzone boundary is an immediate visual check that the points fall within Singapore after projection.

```{r}
plot(st_geometry(mpsz))
plot(st_geometry(preschool), add = TRUE)
```

## Working with Projection

Distance and area calculations should be done in a projected coordinate system. For Singapore, SVY21 / EPSG:3414 is appropriate because its units are metres.

### Assigning EPSG code to a simple feature data frame

```{r}
st_crs(mpsz)
```

```{r}
mpsz <- st_set_crs(mpsz, 3414)
```

```{r}
st_crs(mpsz)
```

### Transforming the projection of preschool from wgs84 to svy21.

```{r}
preschool <- st_transform(preschool, crs = 3414)
```

```{r}
plot(st_geometry(mpsz))
plot(st_geometry(preschool), add = TRUE)
```

::: {.callout-tip}
## What I Noticed

The first overlay makes the preschool points look misplaced because the layers start in different coordinate systems. After transforming to SVY21, the points line up with the subzone boundary, which confirms why CRS checking cannot be skipped.
:::

## Importing and Converting An Aspatial Data

The Airbnb listings file is a CSV table with longitude and latitude columns. It starts as non-spatial data, then becomes an `sf` point layer after assigning WGS84 coordinates and transforming to SVY21.

### Importing the aspatial data

```{r}
listings <- read_csv("data/aspatial/listings.csv")
```

```{r}
#| label: course-chap01-read-listings
#| eval: false
listings <- read_csv("chap01/data/aspatial/listings.csv")
```

```{r}
list(listings)
```

### Creating a simple feature data frame from an aspatial data frame

```{r}
listings_sf <- st_as_sf(listings, 
                        coords = c("longitude", "latitude"), 
                        crs=4326) %>%
  st_transform(crs = 3414)
```

```{r}
glimpse(listings_sf)
```

::: {.callout-tip}
## What I Noticed

This step reminded me that a table with longitude and latitude is not yet a spatial layer until the coordinates and CRS are declared. After conversion, the geometry column becomes part of the data, so later spatial operations can use the Airbnb records in the same way as other point layers.
:::

## Geoprocessing with sf package

### Use case 1: Land acquisition analysis

#### The scenario

This section creates a 5 metre buffer around cycling paths. The result is useful for asking how much land area is covered by a narrow corridor around the cycling network.

#### The solution

```{r}
buffer_cycling <- st_buffer(cyclingpath, 
                            dist=5, 
                            nQuadSegs = 30)
```

```{r}
#| label: course-chap01-buffer-cycling
#| eval: false
buffer_cycling <- st_buffer(
  cyclingpath, dist=5, nQuadSegs = 30)
```

```{r}
buffer_cycling$AREA <- st_area(buffer_cycling)
```

```{r}
buffer_cycling <- buffer_cycling %>%
  mutate(AREA = st_area(geometry))
```

```{r}
sum(buffer_cycling$AREA)
```

The next operation narrows the cycling-path buffer to a single subzone, Tampines West.

```{r}
mpsz_selected <- mpsz %>%
  filter(SUBZONE_N == "TAMPINES WEST")
```

```{r}
buffer_cycling_selected <- st_intersection(buffer_cycling, 
                                           mpsz_selected)
```

```{r}
#| label: course-chap01-buffer-intersection
#| eval: false
buffer_cycling_selected <- st_intersection(
  buffer_cycling, mpsz_selected)
```

```{r}
plot(st_geometry(mpsz_selected), border = "grey30", lwd = 1.5)
plot(st_geometry(buffer_cycling_selected), col = "lightblue", add = TRUE)
```

::: {.callout-note}
## What I Noticed

The cycling path layer is a current local download, so the buffered area should be read as my reproducible result from this dataset.
:::

### Use case 2: To determine the number of pre-schools by planning subzone

#### The scenario

The subzone polygon layer can be used to count how many preschool points fall inside each subzone.

#### The solution

```{r}
mpsz$`PreSch Count`<- lengths(st_intersects(mpsz, preschool))
```

```{r}
summary(mpsz$`PreSch Count`)
```

```{r}
top_n(mpsz, 1, `PreSch Count`)
```

::: {.callout-tip}
## My Finding

The raw count is a first useful ranking, but I would not use it alone for planning interpretation. A large subzone has more room to contain many facilities, so the count needs to be compared with area before saying a place is especially well-served.
:::

## Density Calculation

A count alone can be misleading because subzones have different sizes. Density standardises the count by area, here expressed as preschools per square kilometre.

```{r}
mpsz$Area <- mpsz %>%
  st_area()
```

```{r}
mpsz <- mpsz %>%
  mutate(`PreSch Density` = `PreSch Count`/Area * 1000000)
```

```{r}
hist(mpsz$`PreSch Density`)
```

```{r}
ggplot(data = mpsz,
       aes(x = as.numeric(`PreSch Density`))) +
  geom_histogram(bins = 20,
                 colour = "white",
                 fill = "#0f766e") +
  labs(title = "Distribution of preschool density by subzone",
       x = "Preschools per square kilometre",
       y = "Number of subzones") +
  theme_minimal()
```

```{r}
ggplot(data = st_drop_geometry(mpsz),
       aes(x = `PreSch Count`, 
           y = as.numeric(`PreSch Density`))) +
  geom_point(alpha = 0.65, colour = "#2563eb") +
  labs(title = "Preschool count and density",
       x = "Preschool count",
       y = "Preschools per square kilometre") +
  theme_minimal()
```

```{r}
#| label: course-chap01-preschool-histogram
#| eval: false
ggplot(data=mpsz,
       aes(x= as.numeric(`PreSch Density`)))+
  geom_histogram(bins=20,
                 color="black",
                 fill="light blue") +
  labs(title = "Are pre-school even distributed in Singapore?",
       subtitle= "There are many planning subzones with a single pre-school, on the other hand, \nthere are seven planning subzones with at least 30 or more pre-schools",
      x = "Pre-school density (per km sq)",
      y = "Frequency")
```

```{r}
#| label: course-chap01-preschool-scatter
#| eval: false
ggplot(data=mpsz,
       aes(y = `PreSch Count`,
           x= as.numeric(`PreSch Density`)))+
  geom_point(color="black",
             fill="light blue") +
  xlim(0, 40) +
  ylim(0, 40) +
  labs(title = "",
      x = "Pre-school density (per km sq)",
      y = "Pre-school count")
```

::: {.callout-note}
## My Reading of the Plot

The scatterplot helps me see why count and density should be interpreted together. Some subzones with moderate counts can still have high density, which means the spatial concentration is stronger than the count table alone suggests.
:::

```{r}
top_preschool_count <- mpsz %>%
  st_drop_geometry() %>%
  arrange(desc(`PreSch Count`)) %>%
  select(SUBZONE_N, PLN_AREA_N, `PreSch Count`, `PreSch Density`) %>%
  slice_head(n = 5)

top_preschool_count
```

::: {.callout-tip}
## What I Noticed

With the current preschool layer, Tampines East has the largest raw count in my output, with 72 centres. Cecil has the highest density, about 35.60 centres per square kilometre, from only 7 centres. These two values answer different questions. The count describes total provision. Density describes concentration within the subzone area.
:::

## Takeaway

The workflow shows why CRS checks and geometry operations are central to geospatial analysis. With the current preschool data, the areas with the highest raw counts are not always the same as the areas with the highest density, because compact subzones can have high service concentration even with moderate counts.

## Reflection

The most important lesson for me is that spatial analysis is not only about drawing data on a map. The geometry type, CRS, and spatial unit all affect what question the code is answering. I also realised that a result can be technically correct but still incomplete if I do not explain whether it is a count, density, buffer area, or spatial relationship.

::: {.callout-tip}
## Beyond the Code: CRS Is Not Just a Setting

One detail I would be more careful with after this exercise is the difference between assigning a CRS and transforming a CRS. `st_set_crs()` is only telling R what the coordinates already mean, while `st_transform()` actually recalculates the coordinates into another reference system. If I used the wrong one, the map might still draw, but the distance, area, and density calculations could become misleading.

This also changes how I read the outputs. When I calculate preschool density or spatial intersections, I am not only checking whether the code runs; I am checking whether the geometry is in a coordinate system that matches the question. For Singapore analysis, projected metres are much easier to defend than raw longitude and latitude when area or distance is involved.
:::

## References

- R for Geospatial Data Science and Analytics, Chapter 1: [Geospatial Data Science with R](https://r4gdsa.netlify.app/chap01.html)
- data.gov.sg: [Pre-Schools Location](https://data.gov.sg/datasets/d_61eefab99958fd70e6aab17320a71f1c/view)
- data.gov.sg: [Master Plan 2014 Subzone Boundary No Sea](https://data.gov.sg/datasets/d_226cacceceff94f0c8b814962a5307c9/view)
- LTA DataMall: [Cycling Path geospatial dataset](https://datamall.lta.gov.sg/)
- Inside Airbnb: [Singapore listings data](https://insideairbnb.com/get-the-data/)
