---
title: "Thematic Mapping and GeoVisualisation with R"
subtitle: "Hands-on Exercise 1"
author: "Zhiyi Hao"
date: "2026-08-28"
execute:
  echo: true
  message: false
  warning: false
format:
  html:
    code-fold: false
    number-sections: false
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 6,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

## 2.1 Overview {#overview}

### 2.1.1 Survival Tip {#survival-tip}

I read the function description before changing the map settings, especially for classification and legend arguments.

This page presents my thematic mapping workflow for Singapore's 2024 subzone population data. I prepare the boundary and population table first, then use `tmap` to see how classification, borders, palettes, facets, and outlier highlighting change the story of the same dependency-ratio data.

::: {.callout-note}
**Local implementation note.**

The boundary is stored as GeoJSON with planning-area fields in separate columns. The KML parsing function shows how to obtain those fields from an HTML description; this dataset uses direct column selection.
:::

## 2.2 Getting Started {#getting-started}

```{r}
#| label: ch02-01
pacman::p_load(sf, tmap, tidyverse, rvest)
```

## 2.3 Importing Data into R {#importing-data-into-r}

### 2.3.1 The Data {#the-data}

The boundary layer and the 2024 population table are the two datasets used for the choropleth maps.

### 2.3.2 Importing Geospatial Data into R {#importing-geospatial-data-into-r}

```r
mpsz <- st_read("chap02/data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml") %>%
  st_transform(crs = 3414)
```

```{r}
#| label: ch02-02
mpsz <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaGEOJSON.geojson",
                quiet = TRUE) %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)
```

#### 2.3.2.1 Tidying data {#tidying-data}

```{r}
#| label: ch02-03
extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}
```

The GeoJSON already contains the four attributes. The KML extraction step and the direct selection used here are shown together.

```r
mpsz <- mpsz %>%
  mutate(
    REGION_N = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())
```

```{r}
#| label: ch02-04
mpsz <- mpsz %>%
  select(REGION_N, PLN_AREA_N, SUBZONE_N, SUBZONE_C, geometry)
```

```{r}
#| label: ch02-05
mpsz
```

### 2.3.3 Importing Attribute Data into R {#importing-attribute-data-into-r}

```r
popdata <- read_csv("chap02/data/aspatial/respopagesextod2024.csv")
```

```{r}
#| label: ch02-06
popdata <- read_csv("data/aspatial/respopagesextod2024.csv")
```

The SingStat population table is aspatial. It needs to be summarised by planning area, subzone, and age group before joining to the subzone boundary.

### 2.3.4 Data Preparation {#data-preparation}

#### 2.3.4.1 Data wrangling {#data-wrangling}

```{r}
#| label: ch02-07
popdata2024 <- popdata %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup()%>%
  pivot_wider(names_from=AG, 
              values_from=POP) %>%
  mutate(YOUNG = rowSums(.[3:6])
         +rowSums(.[12])) %>%
mutate(`ECONOMY ACTIVE` = rowSums(.[7:11])+
rowSums(.[13:15]))%>%
mutate(`AGED`=rowSums(.[16:21])) %>%
mutate(`TOTAL`=rowSums(.[3:21])) %>%  
mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
/`ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`, 
       `ECONOMY ACTIVE`, `AGED`, 
       `TOTAL`, `DEPENDENCY`)
```

```{r}
#| label: age-group-check
age_columns <- popdata %>%
  group_by(PA, SZ, AG) %>%
  summarise(POP = sum(Pop), .groups = "drop") %>%
  pivot_wider(names_from = AG, values_from = POP) %>%
  names()
tibble(group = c("YOUNG", "ECONOMY ACTIVE", "AGED"),
       age_columns = c(paste(age_columns[c(3:6, 12)], collapse = ", "),
                       paste(age_columns[c(7:11, 13:15)], collapse = ", "),
                       paste(age_columns[16:21], collapse = ", ")))
```

#### 2.3.4.2 Joining the attribute data and geospatial data {#joining-the-attribute-data-and-geospatial-data}

```{r}
#| label: ch02-08
popdata2024 <- popdata2024 %>%
  mutate_at(.vars = vars(PA, SZ), 
          .funs = list(toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

```{r}
#| label: ch02-09
mpsz_pop2024 <- left_join(mpsz, popdata2024,
                          by = c("SUBZONE_N" = "SZ"))
```

```r
write_rds(mpsz_pop2024, "chap02/data/rds/mpsz_pop2024.rds")
```

```{r}
#| label: ch02-10
write_rds(mpsz_pop2024, "data/rds/mpsz_pop2024.rds")
```

## 2.4 Choropleth Mapping Geospatial Data Using tmap {#choropleth-mapping-geospatial-data-using-tmap}

### 2.4.1 Plotting a choropleth map quickly by using qtm() {#plotting-a-choropleth-map-quickly-by-using-qtm}

```{r}
#| label: ch02-11
tmap_mode("plot")
qtm(shp = mpsz_pop2024, 
    fill = "DEPENDENCY")
```

::: {.callout-note}
## First Impression {.unnumbered}

The quick map is useful as a first check because it confirms that the joined attribute can be mapped immediately. At the same time, it feels too plain for communication: the legend, classification, borders, and layout still need design work before the map can support a clear interpretation.
:::

### 2.4.2 Creating a choropleth map by using tmap’s elements {#creating-a-choropleth-map-by-using-tmaps-elements}

```{r}
#| label: ch02-12
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile", 
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

The next map uses a fuller `tmap` composition: filled polygons, title, border, compass, scale bar, grid, and data credits.

::: {.callout-tip}
## What I Noticed {.unnumbered}

The dependency ratio map is sensitive to classification. I find the quantile map useful for comparison because each class has a similar number of subzones, but it can also make small numeric differences look more dramatic.
:::

The following blocks show how small `tmap` design choices change the same spatial pattern.

#### 2.4.2.1 Drawing a base map {#drawing-a-base-map}

```{r}
#| label: ch02-13
tm_shape(mpsz_pop2024) +
  tm_polygons()
```

#### 2.4.2.2 Drawing a choropleth map using tm_polygons() {#drawing-a-choropleth-map-using-tm_polygons}

```{r}
#| label: ch02-14
tm_shape(mpsz_pop2024)+
  tm_polygons(fill = "DEPENDENCY")
```

#### 2.4.2.3 Drawing a choropleth map using tm_fill() and \*tm_border()\*\* {#drawing-a-choropleth-map-using-tm_fill-and-tm_border}

```{r}
#| label: ch02-15
tm_shape(mpsz_pop2024)+
  tm_fill("DEPENDENCY")
```

```{r}
#| label: ch02-16
tm_shape(mpsz_pop2024)+
  tm_fill("DEPENDENCY") +
  tm_borders()
```

```{r}
#| label: ch02-17
tm_shape(mpsz_pop2024)+
  tm_fill("DEPENDENCY") +
  tm_borders(col = "grey60",
             lwd = 0.1,
             lty = "dashed")
```

### 2.4.3 Data classification methods of tmap {#data-classification-methods-of-tmap}

#### 2.4.3.1 Plotting choropleth maps with built-in classification methods {#plotting-choropleth-maps-with-built-in-classification-methods}

```{r}
#| label: ch02-18
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

```{r}
#| label: ch02-19
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

#### 2.4.3.2 Plotting choropleth map with custome break {#plotting-choropleth-map-with-custome-break}

```{r}
#| label: ch02-20
summary(mpsz_pop2024$DEPENDENCY)
```

```{r}
#| label: ch02-21
tm_shape(mpsz_pop2024)+
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00))) +
  tm_borders(fill_alpha = 0.5)
```

### 2.4.4 Colour Scheme {#colour-scheme}

#### 2.4.4.1 Using ColourBrewer palette {#using-colourbrewer-palette}

```{r}
#| label: ch02-22
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

```{r}
#| label: ch02-23
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "-brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

#### 2.4.4.2 Cartographic Furniture {#cartographic-furniture}

```{r}
#| label: ch02-24
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5)) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg & singstat",
             position = c("left", "bottom"))
```

### 2.4.5 Map Layout {#map-layout}

#### 2.4.5.1 Map Legend {#map-legend}

```{r}
#| label: ch02-25
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_pos_auto_in() +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg & singstat",
             position = c("left", "bottom"))
```

#### 2.4.5.2 Map style {#map-style}

```{r}
#| label: ch02-26
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "-brewer.greens")) + 
  tm_borders(fill_alpha = 0.5) + 
  tmap_style("natural")
```

```{r}
#| label: ch02-27
tmap_style("white")
```

::: {.callout-tip}
## What I Noticed {.unnumbered}

The summary shows that most subzones sit around the middle of the dependency-ratio range, but a few areas are much higher. That is why the fixed break map and the quantile map do not feel the same: one emphasises numeric thresholds, while the other forces spatial comparison by rank.
:::

::: {.callout-tip}
## What I Noticed {.unnumbered}

Mapping young and aged population side by side makes the dependency ratio more understandable. Instead of treating dependency as one abstract number, I can see that it is produced by two different age structures, and those two patterns do not always peak in the same places.
:::

## 2.5 Drawing Small Multiple Choropleth Maps {#drawing-small-multiple-choropleth-maps}

Small multiples make it easier to compare young and aged population patterns without changing datasets between plots.

### 2.5.1 By assigning multiple values to at least one of the aesthetic arguments {#by-assigning-multiple-values-to-at-least-one-of-the-aesthetic-arguments}

```{r}
#| label: ch02-28
#| fig-width: 12
#| fig-height: 8
tm_shape(mpsz_pop2024) + 
  tm_polygons(
    fill = c("YOUNG", "AGED"),
    fill.legend = 
      tm_legend(position = tm_pos_in(
        "right", "bottom")),
    fill.scale = tm_scale_intervals(
      style = "equal", 
      n = 5,
      values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tmap_style("natural")
```

### 2.5.2 By arrange multiples choropleth maps in a grid layout {#by-arrange-multiples-choropleth-maps-in-a-grid-layout}

```{r}
#| label: ch02-29
#| fig-width: 12
#| fig-height: 8
youngmap <- tm_shape(mpsz_pop2024)+ 
  tm_polygons(fill = "YOUNG",
              fill.legend = tm_legend(
                position = tm_pos_in(
                  "right", "bottom"),
                  item.height = 0.8),
              fill.scale = tm_scale_intervals(
                style = "quantile", 
                values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of young population")
                
agedmap <- tm_shape(mpsz_pop2024)+ 
  tm_polygons(fill = "AGED",
              fill.legend = tm_legend(
                position = tm_pos_in(
                  "right", "bottom"),
                item.height = 0.8),
              fill.scale = tm_scale_intervals(
              style = "quantile", 
              values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of aged population")

tmap_arrange(youngmap, agedmap, asp=1, ncol=2)
```

### 2.5.3 By defining a group-by variable in tm_facets() {#by-defining-a-group-by-variable-in-tm_facets}

In tmap 4, `title.size` is a relative text size. I use 1.1 for the regional facet titles so each title fits its panel.

```r
tm_shape(mpsz_pop2024) +
  tm_fill(fill = "DEPENDENCY",
          fill.scale = tm_scale_intervals(
            style = "quantile",
            values = "brewer.blues")) + 
  tm_facets(by = "REGION_N",
            nrow = 2, 
            ncols = 3,
            free.coords=TRUE, 
            drop.units=TRUE) +
  tm_layout(legend.show = TRUE,
            title.position = c("center", "center"), 
            title.size = 20) +
  tm_borders(fill_alpha = 0.5)
```

```{r}
#| label: ch02-30
#| fig-width: 12
#| fig-height: 8
tm_shape(mpsz_pop2024) +
  tm_fill(fill = "DEPENDENCY",
          fill.scale = tm_scale_intervals(
            style = "quantile",
            values = "brewer.blues")) + 
  tm_facets(by = "REGION_N",
            nrow = 2, 
            ncols = 3,
            free.coords=TRUE, 
            drop.units=TRUE) +
  tm_layout(legend.show = TRUE,
            title.position = c("center", "center"), 
            title.size = 1.1) +
  tm_borders(fill_alpha = 0.5)
```

::: {.callout-tip}
## My Reading of the Facets {.unnumbered}

The regional facets are helpful because Singapore-wide maps can make smaller regions hard to compare. With `free.coords = TRUE`, each region gets more visual space, so I can inspect internal variation within regions rather than only noticing the largest polygons on the full island map.
:::

## 2.6 Mappping Spatial Object Meeting a Selection Criterion {#mappping-spatial-object-meeting-a-selection-criterion}

```{r}
#| label: ch02-31
mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION") %>%
  tm_shape() +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile", 
                values = "brewer.greens"),
              fill.legend = tm_legend()) +
  tm_borders(fill_alpha = 0.5)
```

## 2.7 Complementing Thematic Map with Statistical Chart {#complementing-thematic-map-with-statistical-chart}

```{r}
#| label: ch02-32
mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION") %>%
  tm_shape() +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile", 
                values = "brewer.greens"),
              fill.legend = tm_legend(),
              fill.chart = tm_chart_box()) +
  tm_borders() +
  tm_layout(asp = 0.8)
```

```{r}
#| label: ch02-33
mpsz_selected <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")

stats <- boxplot.stats(mpsz_selected$DEPENDENCY)

outlier_vals <- stats$out

outlier_sf <- mpsz_selected[mpsz_selected$DEPENDENCY %in% outlier_vals, ]

tm_shape(mpsz_selected) +
  tm_polygons(fill = "DEPENDENCY",
          fill.scale = tm_scale_intervals(
            style = "quantile", 
            values = "brewer.blues"),
          fill.legend = tm_legend(),
          fill.chart = tm_chart_box()) +
  tm_borders(fill_alpha = 0.5) +
tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2) +
  tm_text("SUBZONE_N", col = "red", size = 0.7) +
  tm_layout(asp = 0.8)
```

## 2.8 Creating Interactive Map {#creating-interactive-map}

```{r}
#| label: ch02-34
#| fig-height: 6
region_selected <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")
region_bbox <- st_bbox(region_selected)

stats <- boxplot.stats(region_selected$DEPENDENCY)
outlier_vals <- stats$out
outlier_sf <- region_selected[region_selected$DEPENDENCY %in% outlier_vals, ]

tmap_mode("view")
tm_shape(region_selected, 
         bbox = region_bbox) +
  tm_fill("DEPENDENCY",
          id = "SUBZONE_N",
          popup.vars = c(
            "Name" = "SUBZONE_N", 
            "Dependency" = "DEPENDENCY")) +
  tm_borders() +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2)
```

```{r}
#| label: ch02-35
tmap_mode("plot")
```

```{r}
#| label: ch02-36
#| fig-height: 6
region_selected <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")
region_bbox <- st_bbox(region_selected)

stats <- boxplot.stats(region_selected$DEPENDENCY)
outlier_vals <- stats$out
outlier_sf <- region_selected[region_selected$DEPENDENCY %in% outlier_vals, ]

tmap_mode("view")

tm_shape(region_selected, 
         bbox = region_bbox) +
  tm_fill("DEPENDENCY",
          id = "SUBZONE_N",
          popup.vars = c(
            "Name" = "SUBZONE_N", 
            "Dependency" = "DEPENDENCY")) +
  tm_borders() +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2) +
  tm_view(set_zoom_limits = c(12,14))
```

```{r}
#| label: ch02-37
tmap_mode("plot")
```

The interactive maps support zooming and inspection of the central-region subzones. Red borders mark the dependency-ratio outliers identified by the boxplot rule.

::: {.callout-note}
## My Finding {.unnumbered}

The central region subset changes the reading scale. On the full-island map, central subzones compete visually with the rest of Singapore; after filtering, the same dependency values become easier to compare locally, and the red outlier borders feel more purposeful.
:::

::: {.callout-tip}
## What I Noticed {.unnumbered}

The central region view makes the outliers easier to inspect than the whole-island map. In my rendered result, places such as Pearl's Hill and Kampong Glam stand out because the dependency ratio is high relative to nearby central subzones.
:::

## Further Exploration and Reflections {.unnumbered}

## Beyond the Basic Choropleth {.unnumbered}

The maps above describe where dependency ratio is high or low. I also wanted to check how much of the story comes from the spatial arrangement itself, not just the distribution of the numbers.

### Same Values, Shuffled Locations {.unnumbered}

```{r}
#| label: ch02-shuffled-dependency-map
set.seed(626)

mpsz_shuffle <- mpsz_pop2024 %>%
  mutate(DEPENDENCY_SHUFFLED = sample(DEPENDENCY))

observed_dependency_map <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(title = "Dependency")) +
  tm_title("Observed dependency ratio") +
  tm_borders(fill_alpha = 0.5)

shuffled_dependency_map <- tm_shape(mpsz_shuffle) +
  tm_polygons(fill = "DEPENDENCY_SHUFFLED",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(title = "Dependency")) +
  tm_title("Same values, shuffled") +
  tm_borders(fill_alpha = 0.5)

tmap_arrange(observed_dependency_map, shuffled_dependency_map, ncol = 2)
```

::: {.callout-important}
## My Finding {.unnumbered}

The shuffled map uses the same dependency-ratio values, but it breaks the connection between value and location. The observed map looks more geographically coherent, while the shuffled map looks noisier. This is a useful reminder that a choropleth is not only about the variable distribution; the spatial arrangement also carries information.
:::

### Visual Area Bias {.unnumbered}

```{r}
#| label: ch02-area-bias-summary
valid_pop_area <- mpsz_pop2024 %>%
  filter(!is.na(TOTAL), TOTAL > 0) %>%
  mutate(area_km2 = as.numeric(st_area(geometry)) / 1000000) %>%
  arrange(TOTAL)

least_pop_half <- valid_pop_area %>%
  slice_head(prop = 0.5)

area_bias_summary <- tibble(
  measure = c(
    "Share of mapped land area in least-populated half of subzones (%)",
    "Share of population in those subzones (%)"
  ),
  value = c(
    round(sum(least_pop_half$area_km2) / sum(valid_pop_area$area_km2) * 100, 1),
    round(sum(least_pop_half$TOTAL) / sum(valid_pop_area$TOTAL) * 100, 1)
  )
)

area_bias_summary
```

::: {.callout-note}
## What This Changes for My Map Reading {.unnumbered}

In my output, the least-populated half of subzones takes up `r area_bias_summary$value[1]`% of the mapped land area but contains only `r area_bias_summary$value[2]`% of the population. That means the eye can easily overweight large, sparsely populated subzones. For population mapping, I should be careful not to read visual area as population importance.
:::

### Does Preschool Provision Broadly Follow Young Population? {.unnumbered}

```{r}
#| label: ch02-preschool-provision-summary
preschool2019 <- st_read("data/geospatial/PreSchoolsLocation.geojson",
                         quiet = TRUE) %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)

preschool_counts <- lengths(st_intersects(mpsz, preschool2019))

preschool_by_sz <- mpsz %>%
  st_drop_geometry() %>%
  mutate(PRESCHOOL_COUNT = preschool_counts) %>%
  select(SUBZONE_N, PRESCHOOL_COUNT)

provision <- mpsz_pop2024 %>%
  st_drop_geometry() %>%
  filter(!is.na(TOTAL), YOUNG > 0) %>%
  left_join(preschool_by_sz, by = "SUBZONE_N") %>%
  mutate(
    PRESCHOOL_COUNT = replace_na(PRESCHOOL_COUNT, 0),
    young_per_preschool = YOUNG / pmax(PRESCHOOL_COUNT, 1),
    has_preschool = PRESCHOOL_COUNT > 0
  )

provision_summary <- provision %>%
  summarise(
    correlation = round(cor(YOUNG, PRESCHOOL_COUNT, use = "complete.obs"), 3),
    subzones_without_preschool = sum(!has_preschool),
    max_young_no_preschool = max(YOUNG[!has_preschool], na.rm = TRUE)
  )

provision_summary
```

```{r}
#| label: ch02-preschool-provision-plot
ggplot(provision,
       aes(x = YOUNG,
           y = PRESCHOOL_COUNT,
           colour = has_preschool)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, colour = "grey30", linewidth = 0.6) +
  scale_colour_manual(values = c("FALSE" = "#c2410c", "TRUE" = "#0f766e"),
                      labels = c("No preschool", "Has preschool"),
                      name = NULL) +
  labs(title = "Preschool count and young population by subzone",
       x = "Residents in the course YOUNG age group",
       y = "Preschools in subzone") +
  theme_minimal()
```

::: {.callout-tip}
## My Finding {.unnumbered}

The correlation between young population and preschool count is `r provision_summary$correlation`, so the broad relationship is positive. However, `r provision_summary$subzones_without_preschool` subzones with young residents have no preschool point in this dataset, and the largest of these still has `r provision_summary$max_young_no_preschool` young residents. I would treat this as an exploratory signal rather than a final provision assessment, because demand also depends on age bands, school capacity, and nearby centres across subzone boundaries.
:::

## Takeaway {.unnumbered}

Different classification and layout choices can change how strongly a choropleth map communicates spatial contrast. For this dependency-ratio example, the same joined dataset supports overview mapping, regional comparison, and focused outlier inspection. I compare the quantile and equal-interval results, then choose a legend and annotations that communicate the same numeric pattern clearly.

## Reflection {.unnumbered}

The extra checks changed how I read the choropleth maps. A map can look persuasive even when the classification, area size, or spatial unit is shaping the visual message. The shuffled map, area-bias table, and preschool-population comparison help me slow down and ask whether the pattern is genuinely spatial, visually exaggerated, or only a first descriptive clue.

::: {.callout-tip}
## Beyond the Code: Classification Is an Argument {.unnumbered}

The part I kept thinking about is that `style = "quantile"`, `style = "jenks"`, and equal intervals can all use the same data but tell slightly different visual stories. Quantile is useful when I want every class to contain a similar number of subzones, but it can exaggerate small numeric differences. Jenks is better at following natural breaks, but the class boundaries are less intuitive to explain without the data beside the map.

So I should not treat the classification argument inside `tm_fill()` as only a design choice. It is part of the analysis. If I were presenting one final map, I would show why that classification was chosen and check whether the main finding still holds when the style changes.
:::

## 2.9 Reference {#reference}

### 2.9.1 All about tmap package {#all-about-tmap-package}

### 2.9.2 Geospatial data wrangling {#geospatial-data-wrangling}

### 2.9.3 Data wrangling {#data-wrangling-1}

- R for Geospatial Data Science and Analytics, Chapter 2: [Thematic Mapping and GeoVisualisation with R](https://r4gdsa.netlify.app/chap02.html)
- data.gov.sg: [Master Plan 2019 Subzone Boundary No Sea](https://data.gov.sg/datasets/d_8594ae9ff96d0c708bc2af633048edfb/view)
- Singapore Department of Statistics: [Geographic Distribution data tables](https://www.singstat.gov.sg/find-data/search-by-theme/population/geographic-distribution/latest-data)
