---
title: "Geospatial Data Science with R"
subtitle: "Hands-on Exercise 1"
author: "Zhiyi Hao"
date: "2026-08-26"
execute:
  echo: true
  message: false
  warning: false
format:
  html:
    code-fold: false
    number-sections: false
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 6,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

## 1.1 Learning Outcome {#learning-outcome}

I use this page to reproduce the Chapter 1 workflow with the local data folder. My focus is on the basic geospatial steps: importing layers, checking CRS, converting tabular coordinates into `sf`, and doing simple geometry operations.

::: {.callout-note}
**Local implementation note.**

All data for this exercise is stored in `Hands-on_Ex/Hands-on_Ex01/data`, so the code below uses local relative paths such as `data/geospatial` and `data/aspatial/listings.csv`.
:::

## 1.2 Data Acquisition {#data-acquisition}

Before running the code, I found it useful to separate the data by spatial type because each type supports a different kind of question.

| Spatial data type | Dataset in this exercise | How I use it |
|---|---|---|
| Points | Preschool locations and Airbnb listings | Represent individual facilities or listings |
| Polygons | Planning subzones | Define administrative areas for counting and density |
| Lines | Cycling paths | Represent a network-like feature for buffering |

### 1.2.1 Extracting the geospatial data sets {#extracting-the-geospatial-data-sets}

::: {.callout-note}
**My learning point.**

The same map can mix points, polygons, and lines, but the operations are not interchangeable. For example, `st_intersects()` answers a point-in-polygon question, while `st_buffer()` expands a line into an area that can be measured.
:::

### 1.2.2 Extracting the aspatial data set {#extracting-the-aspatial-data-set}

The listing table is kept in the local `data/aspatial` folder and is introduced again when it is converted into a spatial layer.

## 1.3 Getting Started {#getting-started}

```{r}
#| label: ch01-01
pacman::p_load(sf, tidyverse)
```

## 1.4 Importing Geospatial Data {#importing-geospatial-data}

The first step is to bring the Singapore planning subzone boundary, cycling path layer, and preschool point layer into R. The Master Plan 2014 boundary and cycling path files are read as vector layers, while the preschool layer is read from the local KML derived from the downloaded GeoJSON.

### 1.4.1 Importing polygon feature data in shapefile format {#importing-polygon-feature-data-in-shapefile-format}

```r
mpsz = st_read(dsn = "chap01/data/geospatial", 
                  layer = "MP14_SUBZONE_WEB_PL")
```

```{r}
#| label: ch01-02
mpsz = st_read(dsn = "data/geospatial", 
                  layer = "MP14_SUBZONE_WEB_PL",
                  quiet = TRUE)
```

### 1.4.2 Importing polyline feature data in shapefile form {#importing-polyline-feature-data-in-shapefile-form}

```r
cyclingpath = st_read(dsn = "chap01/data/geospatial", 
                         layer = "CyclingPathGazette")
```

```{r}
#| label: ch01-03
cyclingpath = st_read(dsn = "data/geospatial", 
                         layer = "CyclingPathGazette",
                         quiet = TRUE)
```

```{r}
#| label: cycling-crs
cyclingpath <- st_set_crs(cyclingpath, 3414)
```

### 1.4.3 Importing GIS data in kml format {#importing-gis-data-in-kml-format}

```r
preschool = st_read("chap01/data/geospatial/PreSchoolsLocation.kml") %>% 
  st_transform(crs = 3414)
```

```{r}
#| label: ch01-04
preschool = st_read("data/geospatial/PreSchoolsLocation.kml", quiet = TRUE) %>% 
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)
```

The geometry column tells us what spatial type is stored in each object. For the planning subzones, the geometry is polygon-based because each row represents an areal unit.

## 1.5 Checking the Content of A Simple Feature Data Frame {#checking-the-content-of-a-simple-feature-data-frame}

### 1.5.1 Working with st_geometry() {#working-with-st_geometry}

```{r}
#| label: ch01-05
st_geometry(mpsz)
```

### 1.5.2 Working with glimpse() {#working-with-glimpse}

```{r}
#| label: ch01-06
glimpse(mpsz)
```

### 1.5.3 Working with head() {#working-with-head}

```{r}
#| label: ch01-07
head(mpsz, n=5)
```

## 1.6 Plotting the Geospatial Data {#plotting-the-geospatial-data}

```{r}
#| label: ch01-08
plot(mpsz)
```

```{r}
#| label: ch01-09
plot(st_geometry(mpsz))
```

```{r}
#| label: ch01-10
plot(mpsz["PLN_AREA_N"])
```

```{r}
#| label: ch01-11
plot(st_geometry(mpsz))
plot(st_geometry(preschool), 
     add = TRUE)
```

The default `plot()` method is a quick diagnostic view. It is not a publication map, but it is useful for confirming that the geometries and selected attributes look sensible.

Overlaying preschool locations on the subzone boundary is an immediate visual check that the points fall within Singapore after projection.

## 1.7 Working with Projection {#working-with-projection}

Distance and area calculations should be done in a projected coordinate system. For Singapore, SVY21 / EPSG:3414 is appropriate because its units are metres.

### 1.7.1 Assigning EPSG code to a simple feature data frame {#assigning-epsg-code-to-a-simple-feature-data-frame}

```{r}
#| label: ch01-12
st_crs(mpsz)
```

```{r}
#| label: ch01-13
mpsz <- st_set_crs(mpsz, 3414)
```

```{r}
#| label: ch01-14
st_crs(mpsz)
```

### 1.7.2 Transforming the projection of preschool from wgs84 to svy21. {#transforming-the-projection-of-preschool-from-wgs84-to-svy21.}

```{r}
#| label: ch01-15
preschool <- st_transform(preschool, 
                              crs = 3414)
```

```{r}
#| label: ch01-16
plot(st_geometry(mpsz))
plot(st_geometry(preschool), add = TRUE)
```

::: {.callout-tip}
## What I Noticed {.unnumbered}

The preschool import already transforms the points to SVY21, so the first overlay is aligned in this run. Repeating `st_transform(..., 3414)` leaves those coordinates unchanged. The subzone assignment step corrects its CRS label while retaining the existing metre coordinates.
:::

## 1.8 Importing and Converting An Aspatial Data {#importing-and-converting-an-aspatial-data}

The Airbnb listings file is a CSV table with longitude and latitude columns. It starts as non-spatial data, then becomes an `sf` point layer after assigning WGS84 coordinates and transforming to SVY21.

### 1.8.1 Importing the aspatial data {#importing-the-aspatial-data}

```r
listings <- read_csv("chap01/data/aspatial/listings.csv")
```

```{r}
#| label: ch01-17
listings <- read_csv("data/aspatial/listings.csv", show_col_types = FALSE)
```

```{r}
#| label: ch01-18
list(listings)
```

### 1.8.2 Creating a simple feature data frame from an aspatial data frame {#creating-a-simple-feature-data-frame-from-an-aspatial-data-frame}

```{r}
#| label: ch01-19
listings_sf <- st_as_sf(listings, 
                       coords = c("longitude", "latitude"),
                       crs=4326) %>%
  st_transform(crs = 3414)
```

```{r}
#| label: ch01-20
glimpse(listings_sf)
```

::: {.callout-tip}
## What I Noticed {.unnumbered}

This step reminded me that a table with longitude and latitude is not yet a spatial layer until the coordinates and CRS are declared. After conversion, the geometry column becomes part of the data, so later spatial operations can use the Airbnb records in the same way as other point layers.
:::

## 1.9 Geoprocessing with sf package {#geoprocessing-with-sf-package}

### 1.9.1 Use case 1: Land acquisition analysis {#use-case-1-land-acquisition-analysis}

#### 1.9.1.1 The scenario {#the-scenario}

This section creates a 5 metre buffer around cycling paths. The result is useful for asking how much land area is covered by a narrow corridor around the cycling network.

#### 1.9.1.2 The solution {#the-solution}

```{r}
#| label: ch01-21
buffer_cycling <- st_buffer(
  cyclingpath, dist=5, nQuadSegs = 30)
```

```{r}
#| label: ch01-22
buffer_cycling$AREA <- st_area(buffer_cycling)
```

```{r}
#| label: ch01-23
buffer_cycling <- buffer_cycling %>%
  mutate(AREA = st_area(geometry))
```

```{r}
#| label: ch01-24
sum(buffer_cycling$AREA)
```

```{r}
#| label: ch01-25
mpsz_selected <- mpsz %>%
  filter(SUBZONE_N == "TAMPINES WEST")
```

```{r}
#| label: ch01-26
buffer_cycling_selected <- st_intersection(
  buffer_cycling, mpsz_selected)
```

The next operation narrows the cycling-path buffer to a single subzone, Tampines West.

::: {.callout-note}
## What I Noticed {.unnumbered}

The cycling path layer is a current local download, so the buffered area should be read as my reproducible result from this dataset.
:::

### 1.9.2 Use case 2: To determine the number of pre-schools by planning subzone {#use-case-2-to-determine-the-number-of-pre-schools-by-planning-subzone}

#### 1.9.2.1 The scenario {#the-scenario-1}

The subzone polygon layer can be used to count how many preschool points fall inside each subzone.

#### 1.9.2.2 The solution {#the-solution-1}

```{r}
#| label: ch01-27
mpsz$`PreSch Count`<- lengths(st_intersects(mpsz, preschool))
```

```{r}
#| label: ch01-28
summary(mpsz$`PreSch Count`)
```

```{r}
#| label: ch01-29
top_n(mpsz, 1, `PreSch Count`)
```

```{r}
#| label: ch01-30
mpsz$Area <- mpsz %>%
  st_area()
```

```{r}
#| label: ch01-31
mpsz <- mpsz %>%
  mutate(`PreSch Density` = `PreSch Count`/Area * 1000000)
```

```{r}
#| label: ch01-32
hist(mpsz$`PreSch Density`)
```

```{r}
#| label: ch01-33
ggplot(data=mpsz, 
       aes(x= as.numeric(`PreSch Density`)))+
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  labs(title = "Are pre-school even distributed in Singapore?",
       subtitle= "There are many planning subzones with a single pre-school, on the other hand, \nthere are seven planning subzones with at least 30 or more pre-schools",
      x = "Pre-school density (per km sq)",
      y = "Frequency")
```

```{r}
#| label: ch01-34
ggplot(data=mpsz, 
       aes(y = `PreSch Count`, 
           x= as.numeric(`PreSch Density`)))+
  geom_point(color="black", 
             fill="light blue") +
  xlim(0, 40) +
  ylim(0, 40) +
  labs(title = "",
      x = "Pre-school density (per km sq)",
      y = "Pre-school count")
```

::: {.callout-tip}
## My Finding {.unnumbered}

The raw count is a first useful ranking, but I would not use it alone for planning interpretation. A large subzone has more room to contain many facilities, so the count needs to be compared with area before saying a place is especially well-served.
:::

## Further Exploration and Reflections {.unnumbered}

```{r}
plot(st_geometry(mpsz_selected), border = "grey30", lwd = 1.5)
plot(st_geometry(buffer_cycling_selected), col = "lightblue", add = TRUE)
```

## Density Calculation {.unnumbered}

A count alone can be misleading because subzones have different sizes. Density standardises the count by area, here expressed as preschools per square kilometre.

```{r}
ggplot(data = mpsz,
       aes(x = as.numeric(`PreSch Density`))) +
  geom_histogram(bins = 20,
                 colour = "white",
                 fill = "#0f766e") +
  labs(title = "Distribution of preschool density by subzone",
       x = "Preschools per square kilometre",
       y = "Number of subzones") +
  theme_minimal()
```

```{r}
ggplot(data = st_drop_geometry(mpsz),
       aes(x = `PreSch Count`, 
           y = as.numeric(`PreSch Density`))) +
  geom_point(alpha = 0.65, colour = "#2563eb") +
  labs(title = "Preschool count and density",
       x = "Preschool count",
       y = "Preschools per square kilometre") +
  theme_minimal()
```

::: {.callout-note}
## My Reading of the Plot {.unnumbered}

The scatterplot helps me see why count and density should be interpreted together. Some subzones with moderate counts can still have high density, which means the spatial concentration is stronger than the count table alone suggests.
:::

```{r}
top_preschool_count <- mpsz %>%
  st_drop_geometry() %>%
  arrange(desc(`PreSch Count`)) %>%
  select(SUBZONE_N, PLN_AREA_N, `PreSch Count`, `PreSch Density`) %>%
  slice_head(n = 5)

top_preschool_count
```

::: {.callout-tip}
## What I Noticed {.unnumbered}

With the current preschool layer, Tampines East has the largest raw count in my output, with 72 centres. Cecil has the highest density, about 35.60 centres per square kilometre, from only 7 centres. These two values answer different questions. The count describes total provision. Density describes concentration within the subzone area.
:::

## Takeaway {.unnumbered}

The workflow shows why CRS checks and geometry operations are central to geospatial analysis. With the current preschool data, the areas with the highest raw counts are not always the same as the areas with the highest density, because compact subzones can have high service concentration even with moderate counts.

## Reflection {.unnumbered}

The most important lesson for me is that spatial analysis is not only about drawing data on a map. The geometry type, CRS, and spatial unit all affect what question the code is answering. I also realised that a result can be technically correct but still incomplete if I do not explain whether it is a count, density, buffer area, or spatial relationship.

::: {.callout-tip}
## Beyond the Code: CRS Is Not Just a Setting {.unnumbered}

One detail I would be more careful with after this exercise is the difference between assigning a CRS and transforming a CRS. `st_set_crs()` is only telling R what the coordinates already mean, while `st_transform()` actually recalculates the coordinates into another reference system. If I used the wrong one, the map might still draw, but the distance, area, and density calculations could become misleading.

This also changes how I read the outputs. When I calculate preschool density or spatial intersections, I am not only checking whether the code runs; I am checking whether the geometry is in a coordinate system that matches the question. For Singapore analysis, projected metres are much easier to defend than raw longitude and latitude when area or distance is involved.
:::

## References {.unnumbered}

- R for Geospatial Data Science and Analytics, Chapter 1: [Geospatial Data Science with R](https://r4gdsa.netlify.app/chap01.html)
- data.gov.sg: [Pre-Schools Location](https://data.gov.sg/datasets/d_61eefab99958fd70e6aab17320a71f1c/view)
- data.gov.sg: [Master Plan 2014 Subzone Boundary No Sea](https://data.gov.sg/datasets/d_226cacceceff94f0c8b814962a5307c9/view)
- LTA DataMall: [Cycling Path geospatial dataset](https://datamall.lta.gov.sg/)
- Inside Airbnb: [Singapore listings data](https://insideairbnb.com/get-the-data/)
