---
title: "Thematic Mapping and GeoVisualisation with R"
subtitle: "Hands-on Exercise 1"
author: "Zhiyi Hao"
date: "2026-08-28"
execute:
  echo: true
  message: false
  warning: false
format:
  html:
    code-fold: false
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 6,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

## Overview

This page presents my thematic mapping workflow for Singapore's 2024 subzone population data. I prepare the boundary and population table first, then use `tmap` to see how classification, borders, palettes, facets, and outlier highlighting change the story of the same dependency-ratio data.

::: {.callout-note}
## Local Implementation Note

The local boundary data is stored as GeoJSON, so the useful planning-area fields are already separate columns. The helper function below is kept for reproducibility when a boundary file contains a `Description` field.
:::

## Installing and Loading the R Packages

```{r}
pacman::p_load(sf, tmap, tidyverse, rvest)
```

## Importing and Preparing the Boundary Layer

```{r}
mpsz <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaGEOJSON.geojson",
                quiet = TRUE) %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)
```

```{r}
extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}
```

```{r}
if ("Description" %in% names(mpsz)) {
  mpsz <- mpsz %>%
    mutate(
      REGION_N = map_chr(Description, extract_kml_field, "REGION_N"),
      PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
      SUBZONE_N = map_chr(Description, extract_kml_field, "SUBZONE_N"),
      SUBZONE_C = map_chr(Description, extract_kml_field, "SUBZONE_C")
    ) %>%
    select(-Name, -Description) %>%
    relocate(geometry, .after = last_col())
} else {
  mpsz <- mpsz %>%
    select(REGION_N, PLN_AREA_N, SUBZONE_N, SUBZONE_C, geometry)
}
```

```{r}
mpsz
```

## Importing and Preparing the Population Data

The SingStat population table is aspatial. It needs to be summarised by planning area, subzone, and age group before joining to the subzone boundary.

```{r}
popdata <- read_csv("data/aspatial/respopagesextod2024.csv")
```

```{r}
young_age <- c("0_to_4", "5_to_9", "10_to_14", "15_to_19")
active_age <- c("20_to_24", "25_to_29", "30_to_34", "35_to_39",
                "40_to_44", "45_to_49", "50_to_54", "55_to_59",
                "60_to_64")
aged_age <- c("65_to_69", "70_to_74", "75_to_79", "80_to_84",
              "85_to_89", "90_and_over")

popdata2024 <- popdata %>%
  group_by(PA, SZ, AG) %>%
  summarise(POP = sum(Pop), .groups = "drop") %>%
  pivot_wider(names_from = AG,
              values_from = POP,
              values_fill = 0) %>%
  mutate(
    YOUNG = rowSums(across(any_of(young_age))),
    `ECONOMY ACTIVE` = rowSums(across(any_of(active_age))),
    AGED = rowSums(across(any_of(aged_age))),
    TOTAL = YOUNG + `ECONOMY ACTIVE` + AGED,
    DEPENDENCY = (YOUNG + AGED) / `ECONOMY ACTIVE`
  ) %>%
  select(PA, SZ, YOUNG, `ECONOMY ACTIVE`, AGED, TOTAL, DEPENDENCY)
```

```{r}
popdata2024 <- popdata2024 %>%
  mutate(PA = toupper(PA),
         SZ = toupper(SZ)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

```{r}
mpsz_pop2024 <- left_join(mpsz, 
                          popdata2024, 
                          by = c("SUBZONE_N" = "SZ"))
```

```{r}
dir.create("data/rds", recursive = TRUE, showWarnings = FALSE)
write_rds(mpsz_pop2024, "data/rds/mpsz_pop2024.rds")
```

## A Quick Choropleth Map

```{r}
tmap_mode("plot")
qtm(shp = mpsz_pop2024, 
    fill = "DEPENDENCY")
```

## Designing the Map

The next map uses a fuller `tmap` composition: filled polygons, title, border, compass, scale bar, grid, and data credits.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile", 
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authority (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

::: {.callout-tip}
## What I Noticed

The dependency ratio map is sensitive to classification. I find the quantile map useful for comparison because each class has a similar number of subzones, but it can also make small numeric differences look more dramatic.
:::

## Choropleth Mapping Choices

The following blocks show how small `tmap` design choices change the same spatial pattern.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons()
```

```{r}
tm_shape(mpsz_pop2024)+
  tm_polygons(fill = "DEPENDENCY")
```

```{r}
tm_shape(mpsz_pop2024)+
  tm_fill("DEPENDENCY")
```

```{r}
tm_shape(mpsz_pop2024)+
  tm_fill("DEPENDENCY") +
  tm_borders()
```

```{r}
tm_shape(mpsz_pop2024)+
  tm_fill("DEPENDENCY") +
  tm_borders(col = "grey60",
             lwd = 0.1,
             lty = "dashed")
```

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

```{r}
summary(mpsz_pop2024$DEPENDENCY)
```

```{r}
tm_shape(mpsz_pop2024)+
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00))) +
  tm_borders(fill_alpha = 0.5)
```

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "-brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5)) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg & singstat",
             position = c("left", "bottom"))
```

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_pos_auto_in() +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg & singstat",
             position = c("left", "bottom"))
```

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "-brewer.greens")) + 
  tm_borders(fill_alpha = 0.5) + 
  tmap_style("natural")
```

```{r}
tmap_style("white")
```

::: {.callout-tip}
## What I Noticed

The summary shows that most subzones sit around the middle of the dependency-ratio range, but a few areas are much higher. That is why the fixed break map and the quantile map do not feel the same: one emphasises numeric thresholds, while the other forces spatial comparison by rank.
:::

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(
    fill = c("YOUNG", "AGED"),
    fill.legend = 
      tm_legend(position = tm_pos_in(
        "right", "bottom")),
    fill.scale = tm_scale_intervals(
      style = "equal", 
      n = 5,
      values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tmap_style("natural")
```

## Small Multiple Maps

Small multiples make it easier to compare young and aged population patterns without changing datasets between plots.

```{r}
youngmap <- tm_shape(mpsz_pop2024)+ 
  tm_polygons(fill = "YOUNG",
              fill.legend = tm_legend(
                position = tm_pos_in(
                  "right", "bottom"),
                  item.height = 0.8),
              fill.scale = tm_scale_intervals(
                style = "quantile", 
                values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of young population")
                
agedmap <- tm_shape(mpsz_pop2024)+ 
  tm_polygons(fill = "AGED",
              fill.legend = tm_legend(
                position = tm_pos_in(
                  "right", "bottom"),
                item.height = 0.8),
              fill.scale = tm_scale_intervals(
              style = "quantile", 
              values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of aged population")

tmap_arrange(youngmap, agedmap, asp=1, ncol=2)
```

```{r}
tm_shape(mpsz_pop2024) +
  tm_fill(fill = "DEPENDENCY",
          fill.scale = tm_scale_intervals(
            style = "quantile",
            values = "brewer.blues")) + 
  tm_facets(by = "REGION_N",
            nrow = 2, 
            ncols = 3,
            free.coords=TRUE, 
            drop.units=TRUE) +
  tm_layout(legend.show = TRUE,
            title.position = c("center", "center"), 
            title.size = 1.1) +
  tm_borders(fill_alpha = 0.5)
```

## Mapping the Central Region

```{r}
mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION") %>%
  tm_shape() +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile", 
                values = "brewer.greens"),
              fill.legend = tm_legend()) +
  tm_borders(fill_alpha = 0.5)
```

```{r}
mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION") %>%
  tm_shape() +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile", 
                values = "brewer.greens"),
              fill.legend = tm_legend(),
              fill.chart = tm_chart_box()) +
  tm_borders() +
  tm_layout(asp = 0.8)
```

```{r}
mpsz_selected <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")

stats <- boxplot.stats(mpsz_selected$DEPENDENCY)

outlier_vals <- stats$out

outlier_sf <- mpsz_selected[mpsz_selected$DEPENDENCY %in% outlier_vals, ]

tm_shape(mpsz_selected) +
  tm_polygons(fill = "DEPENDENCY",
          fill.scale = tm_scale_intervals(
            style = "quantile", 
            values = "brewer.blues"),
          fill.legend = tm_legend(),
          fill.chart = tm_chart_box()) +
  tm_borders(fill_alpha = 0.5) +
tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2) +
  tm_text("SUBZONE_N", col = "red", size = 0.7) +
  tm_layout(asp = 0.8)
```

## Interactive Mapping Code

For the website version, I render this inspection step in plot mode. The important part here is selecting the central region, setting the bounding box, and highlighting outlier subzones.

```{r}
region_selected <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")
region_bbox <- st_bbox(region_selected)

stats <- boxplot.stats(region_selected$DEPENDENCY)
outlier_vals <- stats$out
outlier_sf <- region_selected[region_selected$DEPENDENCY %in% outlier_vals, ]

tmap_mode("plot")

tm_shape(region_selected, 
         bbox = region_bbox) +
  tm_fill("DEPENDENCY",
          id = "SUBZONE_N",
          popup.vars = c(
            "Name" = "SUBZONE_N", 
            "Dependency" = "DEPENDENCY")) +
  tm_borders() +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2) +
  tm_layout(frame = FALSE)
```

```{r}
tmap_mode("plot")
```

```{r}
region_selected <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")
region_bbox <- st_bbox(region_selected)

stats <- boxplot.stats(region_selected$DEPENDENCY)
outlier_vals <- stats$out
outlier_sf <- region_selected[region_selected$DEPENDENCY %in% outlier_vals, ]

tmap_mode("plot")

tm_shape(region_selected, 
         bbox = region_bbox) +
  tm_fill("DEPENDENCY",
          id = "SUBZONE_N",
          popup.vars = c(
            "Name" = "SUBZONE_N", 
            "Dependency" = "DEPENDENCY")) +
  tm_borders() +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2) +
  tm_view(set_zoom_limits = c(12,14))
```

```{r}
tmap_mode("plot")
```

::: {.callout-tip}
## What I Noticed

The central region view makes the outliers easier to inspect than the whole-island map. In my rendered result, places such as Pearl's Hill and Kampong Glam stand out because the dependency ratio is high relative to nearby central subzones.
:::

## Takeaway

Different classification and layout choices can change how strongly a choropleth map communicates spatial contrast. For this dependency-ratio example, the same joined dataset supports overview mapping, regional comparison, and focused outlier inspection. I would use the Jenks or quantile versions for exploration, then refine the legend and annotations depending on the final story.

## References

- R for Geospatial Data Science and Analytics, Chapter 2: [Thematic Mapping and GeoVisualisation with R](https://r4gdsa.netlify.app/chap02.html)
- data.gov.sg: [Master Plan 2019 Subzone Boundary No Sea](https://data.gov.sg/datasets/d_8594ae9ff96d0c708bc2af633048edfb/view)
- Singapore Department of Statistics: [Geographic Distribution data tables](https://www.singstat.gov.sg/find-data/search-by-theme/population/geographic-distribution/latest-data)
