---
title: "Spatio-Temporal Point Patterns Analysis"
subtitle: "Hands-on Exercise 3"
author: "Zhiyi Hao"
date: "2026-09-05"
format:
  html:
    toc: true
    toc-depth: 4
    code-fold: false
    number-offset: 5
execute:
  echo: true
  warning: false
  message: false
---

# Spatio-Temporal Point Patterns Analysis {.course-chapter-title}

```{r}
#| label: setup
#| include: false
options(rgl.useNULL = TRUE)
knitr::opts_chunk$set(fig.width = 8, fig.height = 6, fig.align = "center",
                      warning = FALSE, message = FALSE)
set.seed(626)
```

## Overview

I use the 2023 MODIS fire detections on Bangka Island to explore where and when events concentrate. The overall point map, monthly STKDE, daily STKDE, and space-time K-function each give me a different view of the same observations.

## Learning Outcome

### The research questions

My questions are:

- Are fire locations spatially and spatio-temporally independent?
- Where and when do detections concentrate?
- How do monthly marks, daily marks, and smoothing bandwidths change the interpretation?

## The data

The local files contain a NASA FIRMS 2023 MODIS extract and a geoBoundaries Bangka-Belitung boundary. I extract the largest polygon, Bangka Island, and retain its fire points. This produces 741 detections. The boundary is an ADM1-derived coastline; the course uses a union of sub-district boundaries. Small edge and bandwidth differences can follow from this boundary version.

## Installing and Loading the R packages

I load the packages used for the spatial, temporal, and mapping steps.

```{r}
#| label: ch06-01
pacman::p_load(sf, spatstat, sparr, 
               tmap, stpp, tidyverse,
               animation)
```

## Importing and Preparing Study Area

### Importing study area

The course import paths are followed by the local import. I select the main island before defining the observation window.

```r
kbb <- st_read(dsn="chap06/data/rawdata",
               layer = "Kepulauan_Bangka_Belitung")
```

```{r}
#| label: ch06-02
kbb <- st_read(dsn="data/rawdata",
               layer = "Kepulauan_Bangka_Belitung",
               quiet = TRUE)
kbb_parts <- st_cast(kbb, "POLYGON")
kbb <- kbb_parts[which.max(st_area(kbb_parts)), ]
```

```r
kbb_sf <- st_read(dsn="chap06/data/rawdata",
               layer = "Kepulauan_Bangka_Belitung") %>%
  st_union() %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 32748)
```

```{r}
#| label: ch06-03
kbb_sf <- kbb %>%
  st_union() %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 32748)
```

### Converting OWIN

```{r}
#| label: ch06-04
kbb_owin <- as.owin(kbb_sf)
kbb_owin
```

```{r}
#| label: ch06-05
class(kbb_owin)
```

::: {.callout-tip}
### My observation {.unnumbered}

I transformed the boundary to EPSG:32748 before creating the `owin` object because the next steps use distance and density. Keeping the data in longitude and latitude would make the bandwidth values harder to interpret.
:::

## Importing and Preparing Forest Fire data

```r
fire_sf <- read_csv("chap06/data/rawdata/forestfires.csv") %>%
  st_as_sf(coords = c("longitude", "latitude"),
                       crs = 4326) %>%
  st_transform(crs = 32748)
```

```{r}
#| label: ch06-06
fire_sf <- read_csv("data/rawdata/forestfires.csv", show_col_types = FALSE) %>%
  st_as_sf(coords = c("longitude", "latitude"),
                       crs = 4326) %>%
  st_transform(crs = 32748)
fire_sf <- st_filter(fire_sf, kbb_sf)
```

```{r}
#| label: ch06-07
fire_sf <- fire_sf %>% 
  mutate(DayofYear = yday(acq_date)) %>%
  mutate(Month_num = month(acq_date)) %>%
  mutate(Month_fac = month(acq_date, 
                           label = TRUE, 
                           abbr = FALSE))
```

```{r}
#| label: month-counts
month_counts <- fire_sf %>%
  st_drop_geometry() %>%
  count(Month_fac, name = "detections")
month_counts
```

::: {.callout-tip}
### What I noticed {.unnumbered}

The monthly table contains `r nrow(fire_sf)` detections. The largest monthly count is `r max(month_counts$detections)`, in `r as.character(month_counts$Month_fac[which.max(month_counts$detections)])`. I see a clear concentration in the second half of the year. Linking that pattern to dry weather would require rainfall and satellite-observation coverage data.
:::

## Visualising the Fire Points

### Overall plot

```{r}
#| label: ch06-08
tm_shape(kbb_sf)+
  tm_polygons() +
tm_shape(fire_sf) +
  tm_dots()
```

::: {.callout-tip}
### My observation {.unnumbered}

The points are not spread like a regular grid. Several clusters sit inland, and the edge of the island has fewer detections. This is why I would not stop at a point map: visually dense patches can be affected by symbol overlap, so the density plots below are a better diagnostic.
:::

### Visuaising geographic distribution of forest fires by month

```{r}
#| label: ch06-09
tm_shape(kbb_sf)+
  tm_polygons() +
tm_shape(fire_sf) +
  tm_dots(size = 0.1) +
tm_facets(by="Month_fac", 
            free.coords=FALSE, 
            drop.units = TRUE)
```

::: {.callout-tip}
### What I noticed {.unnumbered}

The monthly facets make the seasonal shift much easier to see. The dense months are not just larger versions of the quiet months; the active areas become more continuous, especially around the second half of the year.
:::

## Computing STKDE by Month

### Extracting forest fires by month

```{r}
#| label: ch06-10
fire_month <- fire_sf %>% 
  select(Month_num)
```

### Creating ppp

```{r}
#| label: ch06-11
fire_month_ppp <- as.ppp(fire_month)
fire_month_ppp
```

```{r}
#| label: ch06-12
summary(fire_month_ppp)
```

```{r}
#| label: ch06-13
any(duplicated(fire_month_ppp))
```

### Including Owin object

```{r}
#| label: ch06-14
fire_month_owin <- fire_month_ppp[kbb_owin]
summary(fire_month_owin)
```

```{r}
#| label: ch06-15
plot(fire_month_owin)
```

::: {.callout-tip}
### My observation {.unnumbered}

The `ppp` summary confirms that the marks are numeric months, which is exactly what `spattemp.density()` expects. The polygon window is also important: it prevents the sea and empty bounding-box space from being treated as part of the observation area.
:::

### Computing Spatio-temporal KDE

I use `sparr::spattemp.density()` to estimate density jointly over space and time.

```{r}
#| label: ch06-16
#| cache: true
st_kde <- spattemp.density(fire_month_owin)
summary(st_kde)
```

::: {.callout-tip}
### My observation {.unnumbered}

The printed bandwidths and density range are calculated from the point pattern above. The temporal bandwidth is in months here. Integer month marks can produce very narrow automatic temporal smoothing; I compare this result with the day-of-year version to see how the time representation affects the surface.
:::

### Plotting the spatio-temporal KDE object

I plot July to December with the common density range and the column-wise panel order used in the exercise.

```{r}
#| label: ch06-17
#| fig-width: 12
#| fig-height: 10
tims <- c(7,8,9,10,11,12)
par(mfcol=c(2,3))
for(i in tims){ 
  plot(st_kde, i, 
       override.par=FALSE, 
       fix.range=TRUE, 
       main=paste("KDE at month",i))
}
```

::: {.callout-tip}
### What I noticed {.unnumbered}

The six panels use `fix.range = TRUE`, which keeps the colour scale common across months. I compare the north-western concentration in July with the southern concentration later in the year. A bright local hotspot and a high monthly total describe different aspects of the pattern. The monthly count table helps me keep those two readings separate.
:::

## Computing STKDE by Day of Year

### Creating ppp object

```{r}
#| label: ch06-18
fire_yday_ppp <- fire_sf %>% 
  select(DayofYear) %>%
  as.ppp()
```

### Including Owin object

```{r}
#| label: ch06-19
fire_yday_owin <- fire_yday_ppp[kbb_owin]
summary(fire_yday_owin)
```

###  

```{r}
#| label: ch06-20
#| cache: true
kde_yday <- spattemp.density(
  fire_yday_owin)
summary(kde_yday)
```

```{r}
#| label: ch06-21
#| fig-show: animate
#| animation.hook: gifski
#| interval: 0.15
#| fig-width: 6
#| fig-height: 6
#| dpi: 90
plot(kde_yday)
```

::: {.callout-tip}
### My observation {.unnumbered}

Daily marks retain changes within each month. I compare the early sparse periods with the late-year concentration and inspect whether hotspots persist across neighbouring dates. The default `plot(kde_yday)` rescales colours between frames, so brightness across this animation does not measure an increase in absolute density. A fixed colour range is needed for that comparison.
:::

## Computing STKDE by Day of Year: Improved method

The bootstrap calculation selects spatial and temporal bandwidths. I then run the specified 9,000-metre and 19-day settings and compare their smoothing with the automatic estimate.

```{r}
#| label: ch06-22
#| cache: true
set.seed(1234)
BOOT.spattemp(fire_yday_owin)
```

### Computing spatio-temporal KDE

```{r}
#| label: ch06-23
#| cache: true
kde_yday <- spattemp.density(fire_yday_owin,
                             h = 9000,
                             lambda = 19)
summary(kde_yday)
```

### Plotting the output spatio-temporal KDE

```{r}
#| label: ch06-24
#| fig-show: animate
#| animation.hook: gifski
#| interval: 0.15
#| fig-width: 6
#| fig-height: 6
#| dpi: 90
plot(kde_yday)
```

::: {.callout-tip}
### What I noticed {.unnumbered}

The bootstrap-guided bandwidths are a useful reminder that space and time need different smoothing scales. A 9000 metre spatial bandwidth is still broad enough to show regional hotspots, while a 19 day temporal bandwidth avoids treating every nearby date as a completely separate pattern.
:::

## Spatio-temporal Point Patterns Analysis: stpp methods

### Preparing spatio-temporal point process object of stpp

```{r}
#| label: ch06-25
coords <- st_coordinates(fire_sf)
```

```{r}
#| label: ch06-26
fire_df <- data.frame(
  x = coords[, 1],  
  y = coords[, 2],
  t = fire_sf$`DayofYear`)
```

```{r}
#| label: ch06-27
coords <- st_coordinates(kbb_sf)
s_region_matrix <- coords[, 1:2]
```

```{r}
#| label: ch06-28
fire_stpp <- as.3dpoints(fire_df)
```

```{r}
#| label: ch06-29
class(fire_stpp)
```

```{r}
#| label: ch06-30
#| fig-width: 10
plot(fire_stpp, s_region_matrix)
```

```{r}
#| label: ch06-31
plot(fire_stpp, s_region_matrix,
     pch = 19, mark = TRUE)
```

::: {.callout-tip}
### My observation {.unnumbered}

Colouring the points by day of year shows that later-year detections occupy many of the dense inland locations. I read this as a space-time concentration, not only a spatial cluster repeated uniformly across the year.
:::

### Computing spatio-temporal k-function

```{r}
#| label: ch06-32
kbb_stik <- STIKhat(fire_stpp,
                    s.region = s_region_matrix,
                    t.region = c(150,250),
                    infectious = TRUE)
```

```{r}
#| label: ch06-33
plotK(kbb_stik, n = 20, 
      L = TRUE, type = "contour")
```

### Guide to interpret the plot

The contour is the transformed space-time K estimate. I read the spatial and temporal axes together, and check the time interval used in the computation. A departure from the theoretical reference is descriptive evidence; a simulation test is needed for a significance claim.

## Further Exploration {.unnumbered}

### Month-by-month spatial density {.unnumbered}

These separate two-dimensional KDEs use a separate maximum for each month. I use them to inspect within-month locations; the joint STKDE above supplies the space-time estimate.

```{r}
#| label: monthly-density-alternative
#| fig-width: 10
#| fig-height: 8
#| fig-cap: "Rendered monthly density surfaces for July to December."
fire_xy <- fire_sf %>%
  mutate(
    x = st_coordinates(.)[, 1],
    y = st_coordinates(.)[, 2]
  ) %>%
  st_drop_geometry()

kbb_outline <- st_as_sf(kbb_sf)

ggplot() +
  geom_sf(data = kbb_outline, fill = "#f8faf5", colour = "#4d5f0c", linewidth = 0.35) +
  stat_density_2d_filled(
    data = filter(fire_xy, Month_num %in% 7:12),
    aes(x = x, y = y, fill = after_stat(level)),
    alpha = 0.72,
    contour_var = "ndensity",
    bins = 7
  ) +
  geom_point(
    data = filter(fire_xy, Month_num %in% 7:12),
    aes(x = x, y = y),
    colour = "#7f1d1d",
    alpha = 0.38,
    size = 0.35
  ) +
  facet_wrap(~ Month_fac, ncol = 3) +
  coord_sf(crs = st_crs(kbb_sf), datum = NA) +
  scale_fill_manual(values = c("#edf8fb", "#ccece6", "#99d8c9", "#66c2a4", "#41ae76", "#238b45", "#005824")) +
  labs(fill = "Relative\ndensity") +
  theme(
    legend.position = "right",
    panel.grid.major = element_line(colour = "#e4ece7", linewidth = 0.25),
    strip.text = element_text(face = "bold")
  )
```

### Broader day-of-year groups {.unnumbered}

Each panel is separately normalised. The grouping shows the locations occupied in each part of the year.

```{r}
#| label: yday-density-alternative
#| fig-width: 10
#| fig-height: 7
#| fig-cap: "Selected day-of-year periods, shown as density and point overlays."
fire_periods <- fire_xy %>%
  mutate(
    period = case_when(
      DayofYear <= 120 ~ "Days 1-120",
      DayofYear <= 210 ~ "Days 121-210",
      DayofYear <= 300 ~ "Days 211-300",
      TRUE ~ "Days 301-365"
    ),
    period = factor(period, levels = c("Days 1-120", "Days 121-210", "Days 211-300", "Days 301-365"))
  )

ggplot() +
  geom_sf(data = kbb_outline, fill = "#fbfbf3", colour = "#60710d", linewidth = 0.35) +
  stat_density_2d_filled(
    data = fire_periods,
    aes(x = x, y = y, fill = after_stat(level)),
    alpha = 0.72,
    contour_var = "ndensity",
    bins = 6
  ) +
  geom_point(
    data = fire_periods,
    aes(x = x, y = y),
    colour = "#831843",
    alpha = 0.35,
    size = 0.32
  ) +
  facet_wrap(~ period, ncol = 2) +
  coord_sf(crs = st_crs(kbb_sf), datum = NA) +
  scale_fill_manual(values = c("#f7fcfd", "#d0ecf0", "#9bd5db", "#66b6c6", "#3288bd", "#0b4f6c")) +
  labs(fill = "Relative\ndensity") +
  theme(
    legend.position = "right",
    panel.grid.major = element_line(colour = "#e2eceb", linewidth = 0.25),
    strip.text = element_text(face = "bold")
  )
```

### Time-marked point map {.unnumbered}

```{r}
#| label: stpp-scatter-alternative
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "Space-time scatter view of fire detections."
ggplot(fire_df, aes(x = x, y = y, colour = t)) +
  geom_point(alpha = 0.58, size = 0.8) +
  geom_sf(data = kbb_outline, inherit.aes = FALSE, fill = NA, colour = "#45520b", linewidth = 0.35) +
  coord_sf(crs = st_crs(kbb_sf), datum = NA) +
  scale_colour_viridis_c(option = "C", name = "Day of year") +
  labs(x = "Easting", y = "Northing") +
  theme(panel.grid.major = element_line(colour = "#e6eeec", linewidth = 0.25))
```

### Nearby event pairs {.unnumbered}

```{r}
#| label: space-time-pair-summary
sp_dist <- as.matrix(dist(as.matrix(fire_df[, c("x", "y")])))
tm_dist <- as.matrix(dist(fire_df$t))
upper_idx <- upper.tri(sp_dist)

pair_summary <- expand_grid(
  spatial_band = c(5000, 10000, 20000),
  temporal_band = c(7, 14, 30)
) %>%
  mutate(
    close_pairs = map2_int(
      spatial_band,
      temporal_band,
      ~ sum(sp_dist[upper_idx] <= .x & tm_dist[upper_idx] <= .y)
    ),
    spatial_label = paste0(spatial_band / 1000, " km"),
    temporal_label = paste0(temporal_band, " days")
  )

pair_summary %>%
  dplyr::select(spatial_label, temporal_label, close_pairs) %>%
  arrange(spatial_label, temporal_label)
```

```{r}
#| label: space-time-pair-heatmap
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "Counts of fire-event pairs within selected spatial and temporal bands."
ggplot(pair_summary, aes(x = temporal_label, y = spatial_label, fill = close_pairs)) +
  geom_tile(colour = "white", linewidth = 0.8) +
  geom_text(aes(label = close_pairs), colour = "white", fontface = "bold") +
  scale_fill_gradient(low = "#a6c936", high = "#7f1d1d") +
  labs(x = "Temporal distance", y = "Spatial distance", fill = "Pairs") +
  theme(
    panel.grid = element_blank(),
    axis.text = element_text(face = "bold")
  )
```

::: {.callout-tip}
### My observation {.unnumbered}

Close-pair counts necessarily increase as either threshold grows because each larger window contains the smaller one. I use these counts as descriptive summaries of proximity. Establishing excess space-time interaction requires a benchmark that preserves the spatial and temporal margins, together with edge correction; raw pair counts alone cannot establish it.
:::

## Final Takeaway {.unnumbered}


This exercise helped me see why spatio-temporal point pattern analysis needs both a spatial window and a time mark. The overall map suggests clustering, but the monthly and day-of-year views show when the clustering becomes strongest. In my local extract, the clearest signal appears after mid-year, with the strongest density around the late dry-season months.

Reproducibility requires the data, study window, method, and reported output to agree. I record the boundary source and run the STKDE and K-function directly. Separate two-dimensional density panels give an additional visual comparison, with their own scale limitations.

::: {.callout-tip}
## Beyond the Code: Time Marks Need a Scale {.unnumbered}

For spatio-temporal KDE, I should not treat `Month_num` and `DayofYear` as interchangeable time marks. Month is easier to communicate and gives a clean seasonal overview, but it hides within-month changes. Day of year is more detailed, but it creates many more time slices and needs stronger smoothing to avoid turning the page into noisy frame-by-frame output.

The `lambda` value is therefore similar to bandwidth in space: it decides how much neighbouring time is allowed to influence a density estimate. In my interpretation, a monthly view is good for telling the broad fire-season story, while the day-of-year view is better for checking whether that story happens gradually or through shorter bursts of activity.
:::



## Reference

- Course source: [Spatio-Temporal Point Patterns Analysis](https://r4gdsa.netlify.app/chap06.html)
- Hall, P. (1990). [Using the Bootstrap to Estimate Mean Squared Error and Select Smoothing Parameter in Nonparametric Problems](https://core.ac.uk/works/48942826).
- NASA FIRMS: [Archive Download](https://firms.modaps.eosdis.nasa.gov/download/).
- [geoBoundaries](https://www.geoboundaries.org/): Indonesia ADM1 boundary.
- [sparr reference](https://tilmandavies.github.io/sparr/reference/spattemp.density.html).
