---
title: "2nd Order Spatial Point Patterns Analysis Methods"
subtitle: "Hands-on Exercise 2"
author: "Zhiyi Hao"
date: "2026-09-02"
format:
  html:
    toc: true
    toc-depth: 4
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 6,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

## Overview

This page continues from the Chapter 4 preparation and focuses on Choa Chu Kang and Tampines. I use G-, F-, K-, and L-functions to compare how childcare centre locations behave across distance.

## The Data

The same two local GeoJSON datasets are used:

- Child Care Services as point features.
- Master Plan 2019 Subzone Boundary (No Sea) as polygon features.

The Chapter 5 analysis focuses on the Choa Chu Kang and Tampines planning areas, using the point pattern objects prepared in the same way as Chapter 4.

## Installing and Loading the R Packages

```{r}
#| label: ch05-packages
pacman::p_load(sf, spatstat, tmap, tidyverse)
```

```{r}
#| label: ch05-seed
set.seed(1234)
```

## Data Import and Preparation

::: {.callout-note}
### Local implementation note

To keep this page reproducible on its own, I recreate the required `sf`, `owin`, and `ppp` objects from the local GeoJSON files.
:::

```{r}
#| label: ch05-read-data
mpsz_cl <- st_read("data/MasterPlan2019SubzoneBoundaryNoSeaGEOJSON.geojson",
                   quiet = TRUE) %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414) %>%
  select(REGION_N, PLN_AREA_N, SUBZONE_N, SUBZONE_C, geometry) %>%
  filter(SUBZONE_N != "SOUTHERN GROUP",
         PLN_AREA_N != "WESTERN ISLANDS",
         PLN_AREA_N != "NORTH-EASTERN ISLANDS")

childcare_sf <- st_read("data/ChildCareServices.geojson",
                        quiet = TRUE) %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)

childcare_ppp <- as.ppp(childcare_sf)
```

```{r}
#| label: ch05-create-planning-patterns
tm <- mpsz_cl %>%
  filter(PLN_AREA_N == "TAMPINES")
ck <- mpsz_cl %>%
  filter(PLN_AREA_N == "CHOA CHU KANG")

tm_owin = as.owin(tm)
ck_owin = as.owin(ck)

childcare_tm_ppp = childcare_ppp[tm_owin]
childcare_ck_ppp = childcare_ppp[ck_owin]
```

```{r}
#| label: ch05-plot-study-areas
par(mfrow=c(1,2))
plot(unmark(childcare_ck_ppp), main = "Choa Chu Kang")
plot(unmark(childcare_tm_ppp), main = "Tampines")
```

## Second-order Spatial Point Patterns Analysis

The observed curve is compared with what complete spatial randomness would look like. When an observed line moves outside the simulation envelope, I read that as evidence that the pattern is behaving differently from CSR for that distance range.

## Analysing Spatial Point Process Using G-Function

::: {.callout-tip}
### How I read this plot

The G-function is event-to-event. I read it as a nearest-neighbour view: if the observed curve rises faster than the CSR expectation, many events have close neighbours.
:::

### Choa Chu Kang Planning Area

#### Computing G-function Estimation

```{r}
#| label: ch05-g-ck
G_CK = Gest(childcare_ck_ppp, correction = "border")
plot(G_CK, xlim=c(0,500))
```

#### Performing Complete Spatial Randomness Test

```{r}
#| label: ch05-g-ck-envelope
G_CK.csr <- envelope(childcare_ck_ppp, Gest, nsim = 999)
```

```{r}
#| label: ch05-plot-g-ck-envelope
plot(G_CK.csr)
```

### Tampines Planning Area

#### Computing G-function Estimation

```{r}
#| label: ch05-g-tm
G_tm = Gest(childcare_tm_ppp, correction = "best")
plot(G_tm)
```

#### Performing Complete Spatial Randomness Test

```{r}
#| label: ch05-g-tm-envelope
G_tm.csr <- envelope(childcare_tm_ppp, Gest, correction = "all", nsim = 999)
```

```{r}
#| label: ch05-plot-g-tm-envelope
plot(G_tm.csr)
```

::: {.callout-tip}
### My finding from the G-function

The G-function is the easiest one for me to connect to nearest-neighbour clustering. In these two planning areas, I read the early part of the curve most carefully because short-distance behaviour tells me whether childcare centres tend to have nearby neighbours more often than expected under CSR.
:::

## Analysing Spatial Point Process Using F-Function

::: {.callout-tip}
### How I read this plot

The F-function is location-to-event. Instead of starting from an existing childcare centre, it starts from arbitrary locations in the study area and asks how far those locations are from the nearest centre.
:::

### Choa Chu Kang Planning Area

#### Computing F-function Estimation

```{r}
#| label: ch05-f-ck
F_CK = Fest(childcare_ck_ppp)
plot(F_CK)
```

### Performing Complete Spatial Randomness Test

```{r}
#| label: ch05-f-ck-envelope
F_CK.csr <- envelope(childcare_ck_ppp, Fest, nsim = 999)
```

```{r}
#| label: ch05-plot-f-ck-envelope
plot(F_CK.csr)
```

### Tampines Planning Area

#### Computing F-function Estimation

```{r}
#| label: ch05-f-tm
F_tm = Fest(childcare_tm_ppp, correction = "best")
plot(F_tm)
```

#### Performing Complete Spatial Randomness Test

```{r}
#| label: ch05-f-tm-envelope
F_tm.csr <- envelope(childcare_tm_ppp, Fest, correction = "all", nsim = 999)
```

```{r}
#| label: ch05-plot-f-tm-envelope
plot(F_tm.csr)
```

::: {.callout-note}
### My finding from the F-function

The F-function adds a different perspective because it starts from empty locations rather than existing centres. If the observed curve rises quickly, it means many locations in the planning area are close to a childcare centre. I find this useful because it connects the point pattern to accessibility-like thinking, even though it is still a geometric summary rather than a service-demand model.
:::

## Analysing Spatial Point Process Using K-Function

::: {.callout-tip}
### How I read this plot

The K-function is more multi-scale. It counts neighbours within increasing distance bands, so it is useful when clustering may happen at some distances but not others.
:::

### Choa Chu Kang Planning Area

#### Computing K-fucntion Estimate

```{r}
#| label: ch05-k-ck
K_ck = Kest(childcare_ck_ppp, correction = "Ripley")
plot(K_ck, . -r ~ r, ylab= "K(d)-r", xlab = "d(m)")
```

#### Performing Complete Spatial Randomness Test

```{r}
#| label: ch05-k-ck-envelope
K_ck.csr <- envelope(childcare_ck_ppp, Kest, nsim = 99, rank = 1, glocal=TRUE)
```

```{r}
#| label: ch05-plot-k-ck-envelope
plot(K_ck.csr, . - r ~ r, xlab="d", ylab="K(d)-r")
```

### Tampines Planning Area

#### Computing K-fucntion Estimation

```{r}
#| label: ch05-k-tm
K_tm = Kest(childcare_tm_ppp, correction = "Ripley")
plot(K_tm, . -r ~ r, 
     ylab= "K(d)-r", xlab = "d(m)", 
     xlim=c(0,1000))
```

#### Performing Complete Spatial Randomness Test

```{r}
#| label: ch05-k-tm-envelope
K_tm.csr <- envelope(childcare_tm_ppp, Kest, nsim = 99, rank = 1, glocal=TRUE)
```

```{r}
#| label: ch05-plot-k-tm-envelope
plot(K_tm.csr, . - r ~ r, 
     xlab="d", ylab="K(d)-r", xlim=c(0,500))
```

::: {.callout-tip}
### My finding from the K-function

The K-function helps me avoid making only a nearest-neighbour statement. I read it as a multi-distance check: clustering can be visible at short distances but become weaker at larger distances, so the distance range matters as much as whether the curve is above or below the CSR reference.
:::

## Analysing Spatial Point Process Using L-Function

::: {.callout-tip}
### How I read this plot

The L-function is a transformed version of K. I find it easier to read because the CSR reference is closer to a flat zero line after plotting `L(d)-r`.
:::

### Choa Chu Kang Planning Area

#### Computing L Fucntion Estimation

```{r}
#| label: ch05-l-ck
L_ck = Lest(childcare_ck_ppp, correction = "Ripley")
plot(L_ck, . -r ~ r, 
     ylab= "L(d)-r", xlab = "d(m)")
```

#### Performing Complete Spatial Randomness Test

```{r}
#| label: ch05-l-ck-envelope
L_ck.csr <- envelope(childcare_ck_ppp, Lest, nsim = 99, rank = 1, glocal=TRUE)
```

```{r}
#| label: ch05-plot-l-ck-envelope
plot(L_ck.csr, . - r ~ r, xlab="d", ylab="L(d)-r")
```

### Tampines Planning Area

#### Computing L-fucntion Estimate

```{r}
#| label: ch05-l-tm
L_tm = Lest(childcare_tm_ppp, correction = "Ripley")
plot(L_tm, . -r ~ r, 
     ylab= "L(d)-r", xlab = "d(m)", 
     xlim=c(0,1000))
```

#### Performing Complete Spatial Randomness Test

```{r}
#| label: ch05-l-tm-envelope
L_tm.csr <- envelope(childcare_tm_ppp, Lest, nsim = 99, rank = 1, glocal=TRUE)
```

```{r}
#| label: ch05-plot-l-tm-envelope
plot(L_tm.csr, . - r ~ r, 
     xlab="d", ylab="L(d)-r", xlim=c(0,500))
```

::: {.callout-note}
### My finding from the L-function

The L-function is helpful as a final check because it makes the K-function easier to read visually. When the transformed curve moves away from the zero reference and outside the simulation envelope, I treat that distance range as stronger evidence of departure from CSR.
:::

## G, F, K and L: How I Distinguish Them

| Function | What it focuses on | Intuition |
|---|---|---|
| G | event-to-event | nearest neighbour |
| F | arbitrary location-to-event | empty space |
| K | neighbours across distance | multi-scale clustering |
| L | transformed K | easier interpretation |

## My Interpretation

Across these plots, I read the observed curves relative to the CSR line and simulation envelope rather than from the curve alone. When a curve sits outside the envelope, it suggests that the childcare centre pattern in that planning area is not behaving like CSR at that distance range. The G and F functions help me think about nearest events and empty space, while K and L give a broader distance-based view.

## What I Learned

- First-order and second-order methods answer different questions: intensity versus interaction.
- Simulation envelopes make the plots more useful because they give a practical CSR comparison.
- The same planning area can look different depending on whether I use G, F, K, or L.
- The interpretation should be tied to distance ranges, not just one overall statement.

## Limitations

CSR is a simple benchmark, and it does not account for population distribution, residential land use, transport access, or planning policy. The result is useful for describing spatial pattern, but not enough by itself to explain the causes of that pattern.

## References

- R for Geospatial Data Science and Analytics. Chapter 5: 2nd Order Spatial Point Patterns Analysis Methods. <https://r4gdsa.netlify.app/chap05>
