---
title: "Local Measures of Spatial Autocorrelation"
subtitle: "Hands-on Exercise 5"
author: "Zhiyi Hao"
date: "2026-09-19"
format:
  html:
    toc: true
    toc-depth: 4
    code-fold: false
    number-offset: 9
    fig-width: 8
    fig-height: 6
execute:
  echo: true
  warning: false
  message: false
---

# Local Measures of Spatial Autocorrelation {.course-chapter-title}

## Overview

This exercise moves from global measures to local measures. I use the same Hunan county data as Exercise 5 Chapter 9, then identify where local clusters, outliers, hot spots, and cold spots occur. The workflow follows the course chapter and keeps the Queen contiguity and distance-weight choices visible.

The main questions are:

- Which counties form significant high-high or low-low GDPPC clusters?
- Which counties behave as spatial outliers compared with their neighbours?
- Do the Getis-Ord Gi* results identify the same broad areas as the local Moran analysis?

## Getting Started

### The analytical question

The global Moran's I in Chapter 9 gave one summary for the whole province. Here I look at the local pattern behind that summary. A significant local statistic identifies a location and its neighbourhood under the selected weights; it does not explain the development mechanism by itself.

### The Study Area and Data

The local folder contains the 88 county-level Hunan boundary units and `Hunan_2012.csv` used in the course example. The boundary is transformed from WGS 84 to UTM zone 50N before distance-based analysis.

### Setting the Analytical Toolls

The original package command is kept visible below. The executable version uses the same package list.

``` r
pacman::p_load(sf, sfdep, tmap, tidyverse)
```

```{r}
#| label: ch10-01
pacman::p_load(sf, sfdep, tmap, tidyverse, knitr)
tmap_mode("plot")
```

## Getting the Data Into R Environment

### Import shapefile into r environment

``` r
hunan <- st_read(dsn = "chap10/data/geospatial",
                 layer = "Hunan") %>%
  st_transform(crs = 32650)
```

```{r}
#| label: ch10-02
hunan <- st_read(
  dsn = "data/geospatial",
  layer = "Hunan",
  quiet = TRUE
) %>%
  st_transform(crs = 32650)
hunan
```

::: callout-note
The course example transforms the Hunan polygons before spatial analysis. Keeping the projected CRS is especially important for the fixed distance band and nearest-neighbour weights later in the chapter.
:::

### Import csv file into r environment

``` r
hunan2012 <- read_csv("chap10/data/aspatial/Hunan_2012.csv")
```

```{r}
#| label: ch10-03
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv", show_col_types = FALSE)
hunan2012 %>% select(County, GDPPC) %>% slice_head(n = 6)
```

### Performing relational join

``` r
hunan <- left_join(hunan,hunan2012) %>%
  select(1:4, 7, 15)
```

```{r}
#| label: ch10-04
hunan <- left_join(hunan, hunan2012, by = "County") %>%
  select(1:4, 7, 15)

tibble(
  county_units = nrow(hunan),
  missing_gdppc = sum(is.na(hunan$GDPPC)),
  mean_gdppc = mean(hunan$GDPPC, na.rm = TRUE)
)
```

### Visualising Regional Development Indicator

``` r
equal <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "GDPPC",
                position = tm_pos_in(
                  "left", "bottom"))) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Equal interval classification")

quantile <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "GDPPC",
                position = tm_pos_in(
                  "left", "bottom"))) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Quantile interval classification")

tmap_arrange(equal,
             quantile,
             asp=1,
             ncol=2)
```

```{r}
#| label: ch10-05
#| fig-width: 12
#| fig-height: 8
equal <- tm_shape(hunan) +
  tm_polygons(
    fill = "GDPPC",
    fill.scale = tm_scale_intervals(
      style = "equal",
      n = 5,
      values = "brewer.blues"),
    fill.legend = tm_legend(
      title = "GDPPC",
      position = tm_pos_in("left", "bottom"))) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Equal interval classification")

quantile <- tm_shape(hunan) +
  tm_polygons(
    fill = "GDPPC",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      n = 5,
      values = "brewer.blues"),
    fill.legend = tm_legend(
      title = "GDPPC",
      position = tm_pos_in("left", "bottom"))) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Quantile interval classification")

tmap_arrange(equal, quantile, asp = 1, ncol = 2)
```

::: callout-tip
### My observation

The equal-interval map emphasises the upper GDPPC tail, while the quantile map makes county-to-county comparison easier. I use both as exploratory views and rely on the local statistics and p-values before calling an area a cluster.
:::

## Local Indicators of Spatial Association (LISA)

### Computing Contiguity Spatial Weights

``` r
wm_q <- hunan %>%
  mutate(nb = st_contiguity(geometry),
         wt = st_weights(
           nb, style = "W"),
         .before = 1)
```

``` r
summary(wm_q$nb)
```

```{r}
#| label: ch10-06
wm_q <- hunan %>%
  mutate(
    nb = st_contiguity(geometry),
    wt = st_weights(nb, style = "W"),
    .before = 1
  )
summary(wm_q$nb)
```

### Row-standardised weights matrix

The `W` style gives each county's neighbours equal total influence. This makes the local scores comparable across counties with different numbers of neighbours.

### Computing local Moran's I

``` r
set.seed(1234)
lisa <- wm_q %>%
  mutate(local_moran = local_moran(
    GDPPC, nb, wt, nsim = 99),
         .before = 1) %>%
unnest(local_moran)
```

``` r
glimpse(lisa)
```

```{r}
#| label: ch10-07
set.seed(1234)
lisa <- wm_q %>%
  mutate(
    local_moran = local_moran(GDPPC, nb, wt, nsim = 99),
    .before = 1
  ) %>%
  unnest(local_moran)

glimpse(lisa)
```

The local Moran output contains the local statistic, expected value, variance, z-score, simulated p-value, and cluster labels. The simulated p-value is the key field for separating a visible contrast from a statistically supported local pattern.

### Mapping local Moran's I values

``` r
tm_shape(lisa) +
  tm_polygons(fill = "ii",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 5,
                values = "brewer.RdBu"),
              fill.legend = tm_legend(
                title = "Local Morans'I",
                position = tm_pos_out())) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Loal Morans'I of GDPPC (Queen's method)")
```

```{r}
#| label: ch10-08
#| fig-width: 8
#| fig-height: 7
tm_shape(lisa) +
  tm_polygons(
    fill = "ii",
    fill.scale = tm_scale_intervals(
      style = "pretty", n = 5, values = "brewer.RdBu"),
    fill.legend = tm_legend(
      title = "Local Moran's I", position = tm_pos_out())) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Local Moran's I of GDPPC (Queen's method)")
```

### Mapping local Moran's I p-values

``` r
tm_shape(lisa) +
  tm_polygons(fill = "p_ii",
              fill.scale = tm_scale_intervals(
                breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
                values = "-brewer.Reds"),
              fill.legend = tm_legend(
                title = "p-value",
      position = tm_pos_out())) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("p-values of Loal Moran's I of GDPPC (Queen's method)")
```

```{r}
#| label: ch10-09
#| fig-width: 8
#| fig-height: 7
tm_shape(lisa) +
  tm_polygons(
    fill = "p_ii",
    fill.scale = tm_scale_intervals(
      breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
      values = "brewer.Reds"),
    fill.legend = tm_legend(
      title = "p-value", position = tm_pos_out())) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("p-values of Local Moran's I of GDPPC (Queen's method)")
```

### Mapping both local Moran's I values and p-values

``` r
ii.map <- tm_shape(lisa) +
  tm_polygons(fill = "ii",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 5,
                values = "brewer.RdBu"),
              fill.legend = tm_legend(
                title = "Local Moran's I",
                position = tm_pos_in(
                  "left", "bottom"))) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Loal Moran's I of GDPPC (Queen's method)")

p_ii.map <- tm_shape(lisa) +
  tm_polygons(fill = "p_ii",
              fill.scale = tm_scale_intervals(
                breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
                values = "-brewer.Reds"),
              fill.legend = tm_legend(
                title = "p-value",
      position = tm_pos_in("left", "bottom")
    )) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("p-values of Loal Moran's I of GDPPC (Queen's method)")

tmap_arrange(ii.map, p_ii.map, asp=1, ncol=2)
```

```{r}
#| label: ch10-10
#| fig-width: 12
#| fig-height: 7
ii.map <- tm_shape(lisa) +
  tm_polygons(
    fill = "ii",
    fill.scale = tm_scale_intervals(
      style = "pretty", n = 5, values = "brewer.RdBu"),
    fill.legend = tm_legend(
      title = "Local Moran's I", position = tm_pos_in("left", "bottom"))) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Local Moran's I of GDPPC (Queen's method)")

p_ii.map <- tm_shape(lisa) +
  tm_polygons(
    fill = "p_ii",
    fill.scale = tm_scale_intervals(
      breaks = c(0, 0.001, 0.01, 0.05, 0.1, 1),
      values = "brewer.Reds"),
    fill.legend = tm_legend(
      title = "p-value", position = tm_pos_in("left", "bottom"))) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("p-values of Local Moran's I (Queen's method)")

tmap_arrange(ii.map, p_ii.map, asp = 1, ncol = 2)
```

::: callout-tip
### What I noticed

The local Moran statistic map shows direction and magnitude, while the p-value map shows where that signal is statistically supported. Reading only the first map would risk turning a strong colour into a cluster claim without checking uncertainty.
:::

## Preparing and Visualising LISA Map

### Plotting Moran scatterplot

``` r
lisa <- lisa %>%
  mutate(lag_GDPPC = st_lag(
    GDPPC, nb, wt),
    .before = 1) %>%
  unnest(lag_GDPPC)
```

``` r
ggplot(data = lisa,
       aes(x = GDPPC,
           y = lag_GDPPC)) +
  geom_point() +
  geom_smooth(method = "lm",
              se = FALSE,
              color = "red") +
  labs(x = "GDPPC",
       y = "Spatial Lag of GDPPC",
       title = "Moran Scatterplot") +
  theme_minimal()
```

``` r
ggplot(data = lisa,
       aes(x = GDPPC,
           y = lag_GDPPC,
           color = mean)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm",
              se = FALSE,
              color = "black") +
  geom_hline(yintercept=mean(lisa$lag_GDPPC), lty=2) +
    geom_vline(xintercept=mean(lisa$GDPPC), lty=2) +
  scale_color_manual(
    values = c("High-High" = "red",
               "Low-Low" = "blue",
               "Low-High" = "lightblue",
               "High-Low" = "pink")) +
  labs(x = "GDPPC",
       y = "Spatial Lag of GDPPC",
       title = "Moran Scatterplot with LISA Quadrants") +
  theme_minimal()
```

```{r}
#| label: ch10-11
#| fig-width: 8
#| fig-height: 6
lisa <- lisa %>%
  mutate(lag_GDPPC = st_lag(GDPPC, nb, wt), .before = 1) %>%
  unnest(lag_GDPPC)

ggplot(lisa, aes(x = GDPPC, y = lag_GDPPC, color = mean)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  geom_hline(yintercept = mean(lisa$lag_GDPPC), linetype = 2) +
  geom_vline(xintercept = mean(lisa$GDPPC), linetype = 2) +
  scale_color_manual(values = c(
    "High-High" = "red", "Low-Low" = "blue",
    "Low-High" = "lightblue", "High-Low" = "pink")) +
  labs(x = "GDPPC", y = "Spatial Lag of GDPPC",
       title = "Moran Scatterplot with LISA Quadrants") +
  theme_minimal()
```

The four quadrants separate high-high and low-low clusters from high-low and low-high outliers. I read the quadrant label together with `p_ii`, because a quadrant describes the relationship while the p-value describes the strength of evidence.

### Plotting Moran scatterplot with standardised variable

``` r
lisa <- lisa %>%
  mutate(z_GDPPC = scale(GDPPC),
         z_lag_GDPPC = scale(lag_GDPPC),
         .before = 1)
```

``` r
ggplot(data = lisa,
       aes(x = z_GDPPC,
           y = z_lag_GDPPC,
           color = mean)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm",
              se = FALSE,
              color = "black") +
  geom_hline(yintercept=mean(lisa$z_lag_GDPPC), lty=2) +
    geom_vline(xintercept=mean(lisa$z_GDPPC), lty=2) +
  scale_color_manual(
    values = c("High-High" = "red",
               "Low-Low" = "blue",
               "Low-High" = "lightblue",
               "High-Low" = "pink")) +
  labs(x = "Standardised GDPPC",
       y = "Standardised Spatial Lag of GDPPC",
       title = "Standardised Moran Scatterplot with LISA Quadrants") +
  theme_minimal()
```

```{r}
#| label: ch10-12
lisa <- lisa %>%
  mutate(
    z_GDPPC = as.vector(scale(GDPPC)),
    z_lag_GDPPC = as.vector(scale(lag_GDPPC)),
    .before = 1
  )

ggplot(lisa, aes(x = z_GDPPC, y = z_lag_GDPPC, color = mean)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  geom_hline(yintercept = mean(lisa$z_lag_GDPPC), linetype = 2) +
  geom_vline(xintercept = mean(lisa$z_GDPPC), linetype = 2) +
  scale_color_manual(values = c(
    "High-High" = "red", "Low-Low" = "blue",
    "Low-High" = "lightblue", "High-Low" = "pink")) +
  labs(x = "Standardised GDPPC",
       y = "Standardised Spatial Lag of GDPPC",
       title = "Standardised Moran Scatterplot with LISA Quadrants") +
  theme_minimal()
```

### Preparing LISA map classes

``` r
signif <- 0.05
```

``` r
lisa <- lisa %>%
  mutate(
    LISA_cluster = ifelse(
      p_ii < 0.05,
      as.character(mean),
      "Insignificant"),
    LISA_cluster = factor(
      LISA_cluster,
      levels = c("Insignificant", "Low-Low", "Low-High", "High-Low", "High-High")
    )
  )
```

```{r}
#| label: ch10-13
signif <- 0.05
lisa <- lisa %>%
  mutate(
    LISA_cluster = if_else(p_ii < signif, as.character(mean), "Insignificant"),
    LISA_cluster = factor(
      LISA_cluster,
      levels = c("Insignificant", "Low-Low", "Low-High", "High-Low", "High-High")
    )
  )
```

### Plotting and visualising LISA map

``` r
lisa_map <- tm_shape(lisa) +
  tm_polygons(
    fill = "LISA_cluster",
    fill.scale = tm_scale_categorical(
      values = c(
        "grey80", "blue", "lightblue", "pink", "red"
      )
    ),
    fill.legend = tm_legend(
      title = "LISA Cluster",
      position = tm_pos_in("left", "bottom"))
  ) +
  tm_borders() +
  tm_title("Local Moran's I Clusters (p < 0.05)")
lisa_map
```

```{r}
#| label: ch10-14
#| fig-width: 9
#| fig-height: 7
lisa_map <- tm_shape(lisa) +
  tm_polygons(
    fill = "LISA_cluster",
    fill.scale = tm_scale_categorical(values = c(
      "grey80", "blue", "lightblue", "pink", "red")),
    fill.legend = tm_legend(
      title = "LISA Cluster", position = tm_pos_in("left", "bottom"))) +
  tm_borders() +
  tm_title("Local Moran's I Clusters (p < 0.05)")
lisa_map
```

::: callout-tip
### My observation

The LISA map is more useful for locating candidate clusters than the global Moran value. I treat grey or insignificant counties as an important result too: a visible high or low GDPPC value is not automatically part of a statistically supported local cluster.
:::

## Hot Spots and Cold Spots Analysis (HCSA)

The Getis-Ord Gi* analysis changes the neighbour definition from shared borders to distance-based proximity. I keep both fixed and adaptive distance weights because the two choices answer slightly different questions about neighbourhood scale.

### Deriving distance weight matrix

#### Computing fixed distance weights

``` r
ct <- critical_threshold(st_geometry(hunan))
ct
```

``` r
hunan_fdw <- hunan %>%
  mutate(nb = include_self(
    st_dist_band(
      st_geometry(geometry),
                 upper = ct)),
    wt = st_weights(nb,
                      style = "W"),
    .before = 1)
```

```{r}
#| label: ch10-15
ct <- critical_threshold(st_geometry(hunan))
ct

hunan_fdw <- hunan %>%
  mutate(
    nb = include_self(st_dist_band(st_geometry(geometry), upper = ct)),
    wt = st_weights(nb, style = "W"),
    .before = 1
  )
```

`critical_threshold()` gives a distance large enough for every county to have at least one neighbour. The `include_self()` step is retained because the Gi* statistic includes the focal county in the local sum.

#### Computing adaptive distance weights

``` r
hunan_adw <- hunan %>%
  mutate(nb = include_self(
    st_knn(
      st_geometry(geometry),
      k = 6)),
    wt = st_weights(
      nb, style = "W"),
    .before = 1)
```

```{r}
#| label: ch10-16
hunan_adw <- hunan %>%
  mutate(
    nb = include_self(st_knn(st_geometry(geometry), k = 6)),
    wt = st_weights(nb, style = "W"),
    .before = 1
  )
```

### Computing local Gi*

``` r
HCSA_fdw <- hunan_fdw %>%
  mutate(
    gistar = local_gstar_perm(
      GDPPC, nb, wt, nsim = 99),
    .before = 1) %>%
  unnest(gistar)
```

```{r}
#| label: ch10-17
set.seed(1234)
HCSA_fdw <- hunan_fdw %>%
  mutate(
    gistar = local_gstar_perm(GDPPC, nb, wt, nsim = 99),
    .before = 1
  ) %>%
  unnest(gistar)
```

### Preparing and Visualising HCSA Map

#### Mapping Gi* with fixed distance weights

``` r
tm_shape(HCSA_fdw) +
  tm_polygons(fill = "gi_star",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 6,
                values = "brewer.rd_bu"),
              fill.legend = tm_legend(
                title = "Gi*",
                position = tm_pos_out())) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Gi* of GDPPC (Fixed Bandwidth d = 60799.91m)")
```

```{r}
#| label: ch10-18
#| fig-width: 8
#| fig-height: 7
tm_shape(HCSA_fdw) +
  tm_polygons(
    fill = "gi_star",
    fill.scale = tm_scale_intervals(
      style = "pretty", n = 6, values = "brewer.rd_bu"),
    fill.legend = tm_legend(title = "Gi*", position = tm_pos_out())) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Gi* of GDPPC (Fixed distance band)")
```

#### Mapping local Gi* p-values

``` r
tm_shape(HCSA_fdw) +
  tm_polygons(fill = "p_sim",
              fill.scale = tm_scale_intervals(
                breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
                values = "-brewer.Reds"),
              fill.legend = tm_legend(
                title = "simulated p-value",
      position = tm_pos_out())) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("p-values of local Gi* of GDPPC (Fixed distance)")
```

```{r}
#| label: ch10-19
#| fig-width: 8
#| fig-height: 7
tm_shape(HCSA_fdw) +
  tm_polygons(
    fill = "p_sim",
    fill.scale = tm_scale_intervals(
      breaks = c(0, 0.001, 0.01, 0.05, 0.1, 1),
      values = "brewer.Reds"),
    fill.legend = tm_legend(
      title = "simulated p-value", position = tm_pos_out())) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("p-values of local Gi* of GDPPC (Fixed distance)")
```

#### Mapping both Gi* values and p-values

``` r
Gi_star_map <- tm_shape(HCSA_fdw) +
  tm_polygons(fill = "gi_star",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 5,
                values = "-brewer.rd_bu"),
              fill.legend = tm_legend(
                title = "local Gi*",
                position = tm_pos_in(
                  "left", "bottom"))) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Local Gi* of GDPPC")

p_values_map <- tm_shape(HCSA_fdw) +
  tm_polygons(fill = "p_sim",
              fill.scale = tm_scale_intervals(
                breaks = c(0, 0.001, 0.01, 0.05, 0.1, 1),
                values = "-brewer.reds"),
              fill.legend = tm_legend(
                title = "p-value",
      position = tm_pos_in("left", "bottom")
    )) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("p-values of local Gi* of GDPPC (fixed distance)")

tmap_arrange(Gi_star_map, p_values_map, asp=1, ncol=2)
```

```{r}
#| label: ch10-20
#| fig-width: 12
#| fig-height: 7
Gi_star_map <- tm_shape(HCSA_fdw) +
  tm_polygons(
    fill = "gi_star",
    fill.scale = tm_scale_intervals(
      style = "pretty", n = 5, values = "brewer.rd_bu"),
    fill.legend = tm_legend(
      title = "local Gi*", position = tm_pos_in("left", "bottom"))) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Local Gi* of GDPPC")

p_values_map <- tm_shape(HCSA_fdw) +
  tm_polygons(
    fill = "p_sim",
    fill.scale = tm_scale_intervals(
      breaks = c(0, 0.001, 0.01, 0.05, 0.1, 1),
      values = "brewer.Reds"),
    fill.legend = tm_legend(
      title = "p-value", position = tm_pos_in("left", "bottom"))) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("p-values of local Gi* (fixed distance)")

tmap_arrange(Gi_star_map, p_values_map, asp = 1, ncol = 2)
```

#### Plotting and visualising HCSA Map

``` r
HCSA_fdw <- HCSA_fdw %>%
  mutate(HCSA_cluster = case_when(
    p_sim > 0.05 ~ "Insignificant",
    p_sim <= 0.05 & cluster == "High" ~ "Hot spot",
    p_sim <= 0.05 & cluster == "Low"  ~ "Cold spot",
    TRUE ~ "Other"),
    HCSA_cluster = factor(
      HCSA_cluster,
      levels = c("Insignificant", "Hot spot", "Cold spot")
    ),
    .before = 1
  )
```

``` r
HCSA_map <- tm_shape(HCSA_fdw) +
  tm_polygons(
    fill = "HCSA_cluster",
    fill.scale = tm_scale_categorical(
      values = c("grey80", "red", "blue")
    ),
    fill.legend = tm_legend(
      title = "HSCA Cluster",
      position = tm_pos_in("left", "bottom"))
  ) +
  tm_borders() +
  tm_title("HCSA Clusters (p < 0.05)")
HCSA_map
```

``` r
tmap_arrange(Gi_star_map, HCSA_map, asp=1, ncol=2)
```

```{r}
#| label: ch10-21
#| fig-width: 12
#| fig-height: 7
HCSA_fdw <- HCSA_fdw %>%
  mutate(
    HCSA_cluster = case_when(
      p_sim > 0.05 ~ "Insignificant",
      p_sim <= 0.05 & cluster == "High" ~ "Hot spot",
      p_sim <= 0.05 & cluster == "Low" ~ "Cold spot",
      TRUE ~ "Other"
    ),
    HCSA_cluster = factor(
      HCSA_cluster,
      levels = c("Insignificant", "Hot spot", "Cold spot")
    ),
    .before = 1
  )

HCSA_map <- tm_shape(HCSA_fdw) +
  tm_polygons(
    fill = "HCSA_cluster",
    fill.scale = tm_scale_categorical(values = c("grey80", "red", "blue")),
    fill.legend = tm_legend(
      title = "HCSA Cluster", position = tm_pos_in("left", "bottom"))) +
  tm_borders() +
  tm_title("HCSA Clusters (p < 0.05)")

tmap_arrange(Gi_star_map, HCSA_map, asp = 1, ncol = 2)
```

::: callout-tip
### My interpretation

LISA describes similarity or contrast between a county and its contiguous neighbours. Gi* instead measures whether high or low values accumulate within a distance-defined neighbourhood. Agreement between the two methods would make a location more convincing as a candidate hot spot; disagreement would show sensitivity to the neighbour definition.
:::

## Summary

The local Moran workflow identifies where the global positive autocorrelation is concentrated and separates clusters from outliers. The Gi* workflow adds a distance-based view of hot and cold spots. Both results depend on the weights matrix, the significance threshold, and the GDPPC variable, so I report the spatial definition with the map rather than treating a colour pattern as self-explanatory.

## References

- [Chapter 10: Local Measures of Spatial Autocorrelation](https://r4gdsa.netlify.app/chap10.html)
- [Chapter 9: Global Measures of Spatial Autocorrelation](https://r4gdsa.netlify.app/chap09.html)
- [sfdep reference documentation](https://sfdep.josiahparry.com/)
