---
title: "Global Measures of Spatial Association"
subtitle: "In-class Exercise 5"
author: "Zhiyi Hao"
format: html
date: "2026-09-26"
date-modified: today
execute:
  echo: true
  warning: false
  message: false
---

# Getting Started

## Setting the Analytical Tools

```{r}
#| label: ic05-01
pacman::p_load(sf, sfdep, tmap, tidyverse)
```

## Getting the Data Into R Environment

### Import shapefile into R environment

```{r}
#| label: ic05-02
hunan <- st_read(dsn = "data/geospatial",
                 layer = "Hunan",
                 quiet = TRUE)
```

### Import csv file into R environment

```{r}
#| label: ic05-03
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv",
                      show_col_types = FALSE)
```

### Performing relational join

```{r}
#| label: ic05-04
hunan_GDPPC <- left_join(hunan, hunan2012) %>%
  select(1:4, 7, 15)
```

## Step 1: Deriving Queen's Contiguity Weights

The code chunk below is used to compute the Queen's contiguity weights.
By default, `st_contiguity(geometry)` uses Queen contiguity. Two polygons are treated as neighbours when they share a boundary or a corner.

```{r}
#| label: ic05-05
wm_q <- hunan_GDPPC %>%
  mutate(
    nb = st_contiguity(geometry),
    wt = st_weights(nb, style = "W"),
    .before = 1
  )
```

::: callout-note
It is a good practice to review the output datatable after the computation is done.
`summary(wm_q$nb)` reports the number of regions, number of non-zero links, link distribution, and whether any region has no neighbour.
:::

```{r}
#| label: ic05-06
summary(wm_q$nb)
glimpse(wm_q)
```

::: callout-note
`st_weights()` requires three inputs:

- `nb`: A neighbour list object created by `st_contiguity()`.
- `style`: Default `"W"` for row-standardised weights. Under `"W"`, each county's neighbour weights sum to 1, making counties with different neighbour counts easier to compare. Other supported coding schemes include `"B"`, `"C"`, `"U"`, `"minmax"`, and `"S"`. `"B"` uses basic binary coding, `"C"` is globally standardised, `"U"` divides `"C"` by the number of neighbours, and `"S"` is a variance-stabilising coding scheme.
- `allow_zero`: If `TRUE`, assigns zero as the lagged value to a zone without neighbours.
:::

## Step 2: Computing Global Moran's I

In the code chunk below, `global_moran()` is used to compute Moran's I value.
The three inputs are the numeric variable `x`, the neighbour list `nb`, and the weight list `wt`.

```{r}
#| label: ic05-07
moranI <- global_moran(wm_q$GDPPC,
                       wm_q$nb,
                       wm_q$wt)

glimpse(moranI)
```

::: callout-note
The printed `I` is about 0.301 for this dataset. A positive Moran's I means neighbouring counties tend to have similar GDPPC values under the selected Queen contiguity and row-standardised weights.
:::

## Performing Global Moran's I Test

Moran's I test is used to assess the statistical evidence for global spatial autocorrelation. With `sfdep`, the test can be performed by using `global_moran_test()`.

```{r}
#| label: ic05-08
moranI_test <- global_moran_test(wm_q$GDPPC,
                                 wm_q$nb,
                                 wm_q$wt)

moranI_test
```

::: callout-note
The default alternative is `"greater"`, so the test checks for positive spatial autocorrelation under randomisation. The report gives the Moran I statistic, expectation, variance, standard deviate, and p-value.
:::

## Performing Global Moran's I Permutation Test

Monte Carlo simulation is used to perform the statistical test in practice. With `sfdep`, it is supported by `global_moran_perm()`.

```{r}
#| label: ic05-09
set.seed(1234)

moranI_perm <- global_moran_perm(wm_q$GDPPC,
                                 wm_q$nb,
                                 wm_q$wt,
                                 nsim = 99)

moranI_perm
```

::: callout-tip
- Use `set.seed()` before simulation to make the computation reproducible.
- The simulation count is read with the observed arrangement, so `nsim = 99` gives 100 arrangements in the report.
- `observed rank` shows where the observed Moran's I sits within the simulated reference distribution, and the p-value summarises how extreme the observed statistic is under random permutation.
:::

## Interpretation

Under Queen contiguity and row-standardised weights, Hunan county GDPPC shows significant positive global spatial autocorrelation. The result should be reported together with the selected neighbour definition, weighting style, and test method.


## Computing local Moran's I
In this section,i learnt how to compute local moran's I of GDPPC at county level by using local_moran() of sfdep package.
```{r}
lisa <- wm_q %>%
  mutate(local_moran = local_moran(
    GDPPC, nb, wt, nsim = 99),
         .before = 1) %>%
  unnest(local_moran)
view(lisa)
```

## Plotting LISA map
We can find three fields contain the LISA categories, they are mean, median and pysal. In general, classification in mean will be used as shown in the code chunk below.
```{r}
lisa_sig <- lisa %>%
  filter(p_ii_sim < 0.05)

tmap_mode("plot")
tm_shape(lisa) +
  tm_polygons() +
  tm_borders(fill_alpha = 0.5) +
tm_shape(lisa_sig) +
  tm_polygons(fill = "mean",
              fill.legend = tm_legend(
                text.size = 1.15,
                width = 6)) +
  tm_borders(fill_alpha = 0.4) +
  tm_layout(legend.title.size = 1.25)
```

```{r}
tmap_mode("plot")
tm_shape(lisa) +
  tm_polygons(fill = "p_ii_sim") +
  tm_borders(fill_alpha = 0.5) +
  tm_title("p-value of local Moran's I") +
  tm_layout(legend.title.size = 1.5)
```

```{r}
wm_idw <- hunan_GDPPC %>%
  mutate(nb = include_self(st_contiguity(geometry)),
         wts = st_inverse_distance(nb, geometry,
                                   scale = 1,
                                   alpha = 1),
         .before = 1)

HCSA <- wm_idw %>%
  mutate(local_Gi = local_gstar_perm(
    GDPPC, nb, wts, nsim = 99),
         .before = 1) %>%
  unnest(local_Gi)
```

```{r}
tmap_mode("plot")
tm_shape(HCSA) +
  tm_polygons(
    fill = "p_sim",
    fill.scale = tm_scale_intervals(
      breaks = c(0, 0.001, 0.01, 0.05, 1.0),
      labels = c("p < 0.001", "p < 0.01", "p < 0.05", "Not Significant"),
      values = c("#de2d26", "#fc9272", "#fee0d2", "#e0e0e0")),
    fill.legend = tm_legend(
      title = "p-values")
  ) +
  tm_title("p-values of G*i for GDP per capita") +
  tm_borders(fill_alpha = 0.5)
```


